### 2019-07-15 ms/ms.tex
stuff

\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}

\title{Notes on observational processes in epidemic models}
\author{Sang Woo Park and Benjamin M Bolker}

\usepackage{tabularx}

\usepackage{amsmath}
\usepackage{natbib}
\usepackage{hyperref}
\bibliographystyle{chicago}
\date{\today}

\usepackage{bm}

\usepackage{afterpage}
\usepackage{pdflscape}

\newcommand{\etal}{\textit{et al.}}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textit{#3}\textbf{]}}}
\newcommand{\bmb}[1]{\comment{cyan}{BMB}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\citetapos}[1]{\citeauthor{#1}'s \citeyearpar{#1}}

\newcommand{\fref}[1]{Fig.~\ref{fig:#1}}
\begin{document}

\maketitle

\section{Introduction}

Mechanistic analyses of an epidemic time series allow us to infer the underlying 
transmission mechanism, estimate biologically relevant parameters, and 
predict the course of an epidemic. In order to make a precise and accurate inference, 
disease modelers often focus on developing more realistic process models. For example,
England and Wales measles time series from the prevaccination era have been analyzed
using many different models -- these model account for time-varying transmission
rates \citep{fine1982measles}, realistic age structure \citep{schenzle1984age},
metapopulation structure \citep{xia2004measles}, continuous-time infection process 
\citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.

Despite the amount of effort put into developing better process models, disease 
modelers often neglect details of the observation process associated with disease
case reports. Most disease models assume that new cases are reported
instantaneously when an individual is infected or shows symptom; one exception is
the model by \citep{he2009plug}, which assumes that infections are counted upon
recovery (because diagnosed cases are put to bed rest and are effectively 
no longer infectious).

Here, we use a simple SIR model to study how observation process affects statistical
inference of underlying parameters. We show that making wrong assumptions about
the observation process can lead to biases in parameter estimates and narrow 
confidence intervals.

\section{Methods}

The Susceptible-Infected-Removed (SIR) model describes how disease spreads in a
homogeneous population:
\begin{equation}
\begin{aligned}
\frac{dS}{dt} &= - \beta S \frac{I}{N}\\
\frac{dI}{dt} &= \beta S \frac{I}{N} - \gamma I\\
\frac{dR}{dt} &= \gamma I,
\end{aligned}
\end{equation}
where $\beta$ is the contact rate, $\gamma$ is the recovery rate, and $N = S + I + R$ is
the total population size. Incidence measured upon infection ($i_1(t)$) can be
defined by the integral:
\begin{equation}
i_1(t) = \int_{t}^{t + \Delta t} \beta S \frac{I}{N} dt,
\end{equation}
where $\Delta t$ is the reporting time step.
Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
new state variable $dC/dt = \beta S I/N$ and take the difference between 
two consecutive reporting periods: $i_1(t) = C(t+\delta t) - C(t)$. Likewise, 
incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
\begin{equation}
i_2(t) = \int_{t}^{t + \Delta t} \gamma I dt,
\end{equation}
or by the difference in cumulative number of recovered cases: $i_2(t) = R(t + \Delta t) - R(t)$.
Finally, we assume that the reported incidence follows a negative binomial 
distribution with mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the 
reporting rate, and a dispersion parameter $\theta$.

\section{Results}

In order to introduce the problem, we first compare the dynamics of $i_1(t)$ and
$i_2(t)$ (Figure ?). Lags in reporting time delay the ``observed'' epidemic peak 
timing and reduce the size of the peak. However, these differences in the reporting
time have small effect to the overall shape of the epidemic curve. In the presence
of observation and process error, we may not expect to be able to distinguish between
the two reporting processes. Note that incidence measured at these two periods are 
different from prevalence, which is represented by the state variable $I(t)$; the
observed dynamics of incidence depend on the reporting time step, whereas those of
prevalence do not (hence we expect incidence and prevalence to be similar only when
the reporting time step is equal to the disease generation time). While it is
uncommon, some models do not make a clear distinction between the two.

Although incidence measured at two different periods are qualitatively very similar,

\bibliography{observation}
\end{document}

### 2019-07-21 ms/ms.tex
stuff

@@
 
-Mechanistic analyses of an epidemic time series allow us to infer the underlying 
+Mechanistic analyses of epidemic time series allow us to infer the underlying 
 transmission mechanism, estimate biologically relevant parameters, and 
-predict the course of an epidemic. In order to make a precise and accurate inference, 
-disease modelers often focus on developing more realistic process models. For example,
-England and Wales measles time series from the prevaccination era have been analyzed
-using many different models -- these model account for time-varying transmission
-rates \citep{fine1982measles}, realistic age structure \citep{schenzle1984age},
+predict the course of an epidemic \citep{breto2009time}. In order to make a 
+precise and accurate inference, disease modelers often focus on developing
+more realistic process models. For example, the same London measles time 
+series from the prevaccination era has been analyzed using multiple models -- 
+these models account for time-varying transmission rates \citep{fine1982measles}, 
+realistic age structure \citep{schenzle1984age},
 metapopulation structure \citep{xia2004measles}, continuous-time infection process 
@@
 
-Despite the amount of effort put into developing better process models, disease 
+Despite the amount of effort put into developing \emph{better} process models, disease 
 modelers often neglect details of the observation process associated with disease
 case reports. Most disease models assume that new cases are reported
-instantaneously when an individual is infected or shows symptom; one exception is
+instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
+kennedy2018modeling, pons2018serotype}) or develops symptom; one exception is
 the model by \citep{he2009plug}, which assumes that infections are counted upon
@@
 distribution with mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the 
-reporting rate, and a dispersion parameter $\theta$.
+reporting rate, and an over dispersion parameter $\theta$.
+
+In this study, we focus on estimating 5 parameters: the basic reproductive
+number $\mathcal R_0 = \beta/\gamma$, mean infectious period $1/\gamma$, 
+reporting rate $\rho$, an over dispersion parameter $\theta$ and the initial
+proportion of the infected individuals $i_0$. Initial proportion of
+susceptible individuals is assumed to be $1 - i_0$. The total population
+size $N$ is assumed to be known.
 
@@
 
-Although incidence measured at two different periods are qualitatively very similar,
+Although the observed epidemic curves look very similar regardless of when incidence
+is reported, fitting $i_1(t)$ and $i_2(t)$ to the same time series yields different
+estimates (Figure ?).
+
+To better understand how assumptions about timing of case reporting affect parameter
+estimates of the SIR model, we simulate epidemic curves, measured upon both infection and
+recovery, 100 times and fit both models to each time series (Figure ?).
 

### 2019-07-21 R/fit.R

@@ -36,6 +36,7 @@ fit_sir <- function(data,
 					start=c(log.R0=log(2), log.gamma=log(1), logit.I0=qlogis(1e-4), log.size=log(10),
 							logit.rho=qlogis(0.5)),
 					N=1e5,
+					fixed=NULL,
 					type=c("incidence", "mortality"),
 					tlength=0.1) {
 	type <- match.arg(type)
@@ -44,11 +45,12 @@ fit_sir <- function(data,
 		nll_sir,
 		start,
 		"Nelder-Mead",
+		fixed=fixed,
 		data=list(data=data, tlength=tlength, type=type, N=N),
 		control=list(maxit=5000)
 	)
 	
-	cc <- bbmle::confint(bb, 1)
+	cc <- confint(bb, 1)
 	
 	while (class(cc)=="mle2") {
 		new_start <- coef(cc)
@@ -58,11 +60,12 @@ fit_sir <- function(data,
 			nll_sir,
 			new_start,
 			"Nelder-Mead",
+			fixed=fixed,
 			data=list(data=data, tlength=tlength, type=type, N=N),
 			control=list(maxit=5000)
 		)		
 		
-		cc <- bbmle::confint(bb, 1)
+		cc <- confint(bb, 1)
 	}
 	
 	list(

### 2019-07-21 analysis/incidence_incidence.R

@@ -12,6 +12,8 @@ for (i in 1:length(incidence_data)) {
 	print(i)
 	data <- incidence_data[[i]]
 	
+	fixed <- NULL
+	
 	ff <- fit_sir(data)
 	
 	reslist[[i]] <- ff

### 2019-07-21 analysis/incidence_incidence_fixed.R

@@ -12,11 +12,11 @@ for (i in 1:length(incidence_data)) {
 	print(i)
 	data <- incidence_data[[i]]
 	
-	fixed <- NULL
+	fixed <- c(log.gamma=0)
 	
-	ff <- fit_sir(data, type="mortality")
+	ff <- fit_sir(data, fixed=fixed)
 	
 	reslist[[i]] <- ff
 }
 
-save("reslist", file="incidence_mortality.rda")
+save("reslist", file="incidence_incidence_fixed.rda")

### 2019-07-21 analysis/incidence_mortality_fixed.R

@@ -12,11 +12,11 @@ for (i in 1:length(incidence_data)) {
 	print(i)
 	data <- incidence_data[[i]]
 	
-	fixed <- NULL
+	fixed <- c(log.gamma=0)
 	
-	ff <- fit_sir(data, type="mortality")
+	ff <- fit_sir(data, type="mortality", fixed=fixed)
 	
 	reslist[[i]] <- ff
 }
 
-save("reslist", file="incidence_mortality.rda")
+save("reslist", file="incidence_mortality_fixed.rda")

### 2019-07-21 analysis/mortality_incidence.R

@@ -12,6 +12,8 @@ for (i in 1:length(mortality_data)) {
 	print(i)
 	data <- mortality_data[[i]]
 	
+	fixed <- NULL
+	
 	ff <- fit_sir(data)
 	
 	reslist[[i]] <- ff

### 2019-07-21 analysis/mortality_incidence_fixed.R

@@ -12,11 +12,11 @@ for (i in 1:length(mortality_data)) {
 	print(i)
 	data <- mortality_data[[i]]
 	
-	fixed <- NULL
+	fixed <- c(log.gamma=0)
 	
-	ff <- fit_sir(data, type="mortality")
+	ff <- fit_sir(data, fixed=fixed)
 	
 	reslist[[i]] <- ff
 }
 
-save("reslist", file="mortality_mortality.rda")
+save("reslist", file="mortality_incidence_fixed.rda")

### 2019-07-21 analysis/mortality_mortality_fixed.R

@@ -12,11 +12,11 @@ for (i in 1:length(mortality_data)) {
 	print(i)
 	data <- mortality_data[[i]]
 	
-	fixed <- NULL
+	fixed <- c(log.gamma=0)
 	
-	ff <- fit_sir(data, type="mortality")
+	ff <- fit_sir(data, type="mortality", fixed=fixed)
 	
 	reslist[[i]] <- ff
 }
 
-save("reslist", file="mortality_mortality.rda")
+save("reslist", file="mortality_mortality_fixed.rda")

### 2019-07-21 figure/compare_deterministic.R

@@ -50,19 +50,10 @@ g1 <- ggplot(summdata) +
 	)
 
 g2 <- ggplot(summdata) +
-	geom_boxplot(aes(x=data, y=growth, fill=fit), width=0.5) +
+	geom_boxplot(aes(x=data, y=period, fill=fit), width=0.5) +
 	geom_hline(yintercept=1, lty=2) +
 	scale_x_discrete("Simulations") +
-	scale_y_continuous(expression(Growth~rate)) +
-	theme(
-		legend.position="none"
-	)
-
-g3 <- ggplot(summdata) +
-	geom_boxplot(aes(x=data, y=rho, fill=fit), width=0.5) +
-	geom_hline(yintercept=0.5, lty=2) +
-	scale_x_discrete("Simulations") +
-	scale_y_continuous("Reporting rate") +
+	scale_y_continuous("Mean infectious period") +
 	theme(
 		legend.position="none"
 	)
@@ -73,7 +64,7 @@ coverdata <- summdata %>%
 		coverage=mean(coverage)
 	)
 
-g4 <- ggplot(coverdata) +
+g3 <- ggplot(coverdata) +
 	geom_rect(xmin=-Inf, xmax=Inf, ymin=binom.test(95, 100)[[4]][1], ymax=binom.test(95, 100)[[4]][2], alpha=0.1) +
 	geom_hline(yintercept=0.95, lty=2) +
 	geom_point(aes(x=data, y=coverage, col=fit, group=fit), size=5, position = position_dodge(width = 0.5)) +
@@ -84,6 +75,6 @@ g4 <- ggplot(coverdata) +
 		legend.position=c(0.76, 0.2)
 	)
 
-gtot <- arrangeGrob(g1, g2, g3, g4, nrow=2)
+gtot <- arrangeGrob(g1, g2, g3, nrow=1)
 
-ggsave("compare_deterministic.pdf", gtot, width=6, height=6)
+ggsave("compare_deterministic.pdf", gtot, width=8, height=3)

### 2019-07-22 ms/ms.tex
almost done?

@@
 
-\title{Notes on observational processes in epidemic models}
-\author{Sang Woo Park and Benjamin M Bolker}
+\title{A note on observational processes in epidemic models}
+\author{Sang Woo Park and Benjamin M. Bolker}
 
@@
 
-Here, we use a simple SIR model to study how observation process affects statistical
-inference of underlying parameters. We show that making wrong assumptions about
-the observation process can lead to biases in parameter estimates and narrow 
+Here, we use a simple Susceptible-Infected-Removed (SIR) model to study how 
+assumptions about the underlying observation process affect parameter estimates
+of the SIR model. We show that making wrong assumptions about the timing of 
+incidence reports can lead to biases in parameter estimates and narrow 
 confidence intervals.
@@
 where $\beta$ is the contact rate, $\gamma$ is the recovery rate, and $N = S + I + R$ is
-the total population size. Incidence measured upon infection ($i_1(t)$) can be
-defined by the integral:
+the total population size. We define incidence at time $t$ as the number of newly infected
+individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
+the reporting time step. We expect infected individuals to report their infection some time
+after their infection; the number of reported cases defines the \emph{observed} incidence.
+
+For brevity, we consider two extreme cases: individuals instantaneously report
+their infection when they become infected or when they recover. The observed incidence 
+measured upon infection ($i_1(t)$) can be defined by the integral:
 \begin{equation}
-i_1(t) = \int_{t}^{t + \Delta t} \beta S \frac{I}{N} dt,
+i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt,
 \end{equation}
@@
 Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
-new state variable $dC/dt = \beta S I/N$ and take the difference between 
-two consecutive reporting periods: $i_1(t) = C(t+\delta t) - C(t)$. Likewise, 
+new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
+the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
 incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
 \begin{equation}
-i_2(t) = \int_{t}^{t + \Delta t} \gamma I dt,
+i_2(t) = \int_{t-\Delta t}^{t} \gamma I dt,
 \end{equation}
-or by the difference in cumulative number of recovered cases: $i_2(t) = R(t + \Delta t) - R(t)$.
-Finally, we assume that the reported incidence follows a negative binomial 
-distribution with mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the 
-reporting rate, and an over dispersion parameter $\theta$.
+or by the consecutive difference in the cumulative number of recovered cases:
+$i_2(t) = R(t) - R(t - \Delta t)$.
+Finally, we model under-reporting using a negative binomial distribution with
+mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
+an over dispersion parameter $\theta$. For convenience, we will refer to these two
+negative binomial models as infection model and recovery model hereafter; 
+likewise, we will refer to epidemic time series generated from these models 
+as infection time series and recovery time series.
 
@@
 
-In order to introduce the problem, we first compare the dynamics of $i_1(t)$ and
-$i_2(t)$ (Figure ?). Lags in reporting time delay the ``observed'' epidemic peak 
-timing and reduce the size of the peak. However, these differences in the reporting
-time have small effect to the overall shape of the epidemic curve. In the presence
-of observation and process error, we may not expect to be able to distinguish between
-the two reporting processes. Note that incidence measured at these two periods are 
-different from prevalence, which is represented by the state variable $I(t)$; the
-observed dynamics of incidence depend on the reporting time step, whereas those of
-prevalence do not (hence we expect incidence and prevalence to be similar only when
-the reporting time step is equal to the disease generation time). While it is
+In order to introduce the problem, we first compare the dynamics of observed 
+incidence, $i_1(t)$ and $i_2(t)$, (Figure 1). Lags in reporting time delay 
+the observed epidemic peak timing and reduce the size of the peak. However, 
+these differences in the reporting time have small effect to the overall shape 
+of the epidemic curve. In the presence of observation and process error, we 
+would not expect to be able to distinguish between the two reporting processes.
+
+Note that incidence measured at these two periods are different from prevalence,
+$I(t)$, which is defined as the number of \emph{currently} infected individuals;
+the dynamics of observed incidence depend on the reporting time step (because the sum of 
+true incidence is equal to the final size of an epidemic), whereas those of
+prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
+the reporting time step is equal to the disease generation time. While it is
 uncommon, some models do not make a clear distinction between the two.
 
-Although the observed epidemic curves look very similar regardless of when incidence
-is reported, fitting $i_1(t)$ and $i_2(t)$ to the same time series yields different
-estimates (Figure ?).
+\begin{figure}
+\includegraphics[width=\textwidth]{../figure/example.pdf}
+\caption{
+\textbf{A comparison of incidence measured at two different time points in infection.}
+We used the following parameters to simulate the SIR model and generate the figure: 
+$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
+and $\Delta t = 0.1$ time units.
+}
+\end{figure}
 
-To better understand how assumptions about timing of case reporting affect parameter
-estimates of the SIR model, we simulate epidemic curves, measured upon both infection and
-recovery, 100 times and fit both models to each time series (Figure ?).
+In order to understand how assumptions about the timing of case reporting affect 
+parameter estimates of the SIR model, we simulate observed epidemic curves 
+(i.e., infection time series and recovery time series), measured upon both infection and
+recovery, 100 times and fit both models (i.e., infection model and recovery model) to each 
+time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
+and its coverage probability, defined as the proportion of confidence intervals that
+contain the true value (95\% confidence interval is expected to contain the true value
+95\% of the time by definition). We summarize the results in Figure 2.
+
+When we try to estimate all 5 parameters, fitting the recovery model to
+infection time series underestimates the basic reproduction number and 
+gives a slightly low coverage (Figure 2A). Fitting the infection model to recovery
+time series slightly overestimates the basic reproduction number but gives 
+good coverage. Fitting the correct models to their corresponding time series
+gives unbiased estimates and good coverage.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{../figure/compare_deterministic.pdf}
+\caption{
+\textbf{A comparison of incidence measured at two different time points in infection.}
+We used the following parameters to simulate the SIR model: 
+$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
+$\theta = 10$, and $\Delta t = 0.1$ time units.
+}
+\end{figure}
+
+Disease modelers often assume that the mean infectious period of a disease
+is known and focus on estimating the basic reproduction number (or the 
+transmission rate). When we assume that the mean infectious period is known
+and try to estimate the remaining 4 parameters of the SIR model, fitting the
+wrong model results in clearer bias and lower coverage (Figure 2B); however, 
+bias from fitting the wrong model changes direction (e.g., fitting the infection
+model to recovery model overestimates the reproduction number instead).
+
+Differences in the direction of the bias can be explained by the estimates of the
+exponential growth rate ($r = \beta - \gamma$) and the mean infectious period ($1/\gamma$).
+When we fit the recovery model to infection time series, the model overestimates
+the initial growth rate in order to match the higher epidemic peak of the 
+infection time series. When the mean infectious period is fixed, higher growth rate
+translates to higher basic reproduction number as we see in Figure 2B. When we allow
+the mean infectious period to vary, the model underestimates the mean infectious
+period (high $\gamma$), which also decreases the estimate of the basic reproduction number.
+Likewise, fitting the infection model to recovery time series overestimates the
+mean infectious period (low $\gamma$), increasing the estimate of the basic reproduction number;
+when the mean infectious period is fixed, low growth rate estimates (to match the
+lower epidemic peak of the recovery time series) decrease the estimate of the
+basic reproduction number.
+
+\section{Discussion}
+
+Mathematical modeling of infectious disease outbreaks helps us understand 
+how disease spreads in a population. Disease modelers often put a lot of
+effort into capturing the transmission mechanism but make naive assumptions
+about how cases are reported. Here, we used a simple deterministic SIR model to show that 
+making wrong assumptions about observational processes in epidemic models
+can give biased estimates of the basic reproduction number and narrow
+confidence intervals.
+
+We assumed that newly infected cases are reported instantaneously when
+individuals become infected or when they recover. Although neither of these
+assumptions are realistic, they are broadly consistent with the assumptions
+of other state-of-art epidemic models. More realistic models may distinguish
+reported and unreported cases by adding a new state variable or by modeling
+an explicit delay kernel in reporting time.
+
+While we have only considered observation processes associated with incidence 
+reporting in this study, it is equally as important to consider observation
+processes in mortality reporting. In particular, many disease modelers have
+tried to infer underlying transmission mechanisms from historical mortality 
+data but assumed that deaths are recorded as soon as individuals die. This assumption
+does not account for the possibility that delays in reporting time of deaths can
+change the shape of an epidemic curve: longer delays are expected to decrease
+the size of an epidemic peak and delay the timing of the peak (Figure 1). 
+These assumptions may have important effects on their conclusions about the underlying 
+transmission mechanism.
+
+Our study shows that seemingly irrelevant changes in the assumptions of an epidemic
+model affects the inference of infectious disease transmission.
+We caution disease modelers to be more mindful about their decisions in developing
+epidemic models and the implications of their model assumptions.
 

### 2019-07-22 figure/compare_deterministic.R

@@ -1,6 +1,7 @@
 library(dplyr)
 library(ggplot2); theme_set(theme_bw())
 library(gridExtra)
+library(grid)
 
 scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
@@ -9,24 +10,30 @@ dir <- "../analysis/"
 files <- c("incidence_incidence.rda",
 		   "incidence_mortality.rda",
 		   "mortality_incidence.rda",
-		   "mortality_mortality.rda")
+		   "mortality_mortality.rda",
+		   "incidence_incidence_fixed.rda",
+		   "incidence_mortality_fixed.rda",
+		   "mortality_incidence_fixed.rda",
+		   "mortality_mortality_fixed.rda")
 
-data <- c("infection", "infection", "recovery", "recovery")
-fit <- c("infection", "recovery", "infection", "recovery")
+data <- rep(c("infection", "infection", "recovery", "recovery"), 2)
+fit <- rep(c("infection", "recovery", "infection", "recovery"), 2)
+inf <- rep(c("estimated", "fixed"), each=4)
 
-comblist <- summlist <- vector('list', 4)
+comblist <- summlist <- vector('list', 8)
 
-for (i in 1:4) {
+for (i in 1:8) {
 	load(paste0(dir, files[i]))
 	
 	comblist[[i]] <- reslist
 }
 
-for (i in 1:4) {
+for (i in 1:8) {
 	summlist[[i]] <- lapply(comblist[[i]], function(x) {
 		data.frame(
 			data=data[i],
 			fit=fit[i],
+			inf=inf[i],
 			estimate=x[[1]],
 			growth=(exp(coef(x[[3]])[[1]])-1) * exp(coef(x[[3]])[[2]]),
 			period=1/exp(coef(x[[3]])[[2]]),
@@ -40,41 +47,40 @@ for (i in 1:4) {
 summdata <- summlist %>%
 	bind_rows
 
-g1 <- ggplot(summdata) +
+g1 <- ggplot(filter(summdata, inf=="estimated")) +
 	geom_boxplot(aes(x=data, y=estimate, fill=fit), width=0.5) +
 	geom_hline(yintercept=2, lty=2) +
-	scale_x_discrete("Simulations") +
[…]

### 2019-07-22 figure/example.R

@@ -1,0 +1,32 @@
+library(ggplot2); theme_set(theme_bw())
+library(tidyr)
+library(dplyr)
+library(deSolve)
+source("../R/model.R")
+source("../R/run.R")
+
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+rr <- run_sir()
+
+rr_data <- rr %>%
+	gather(key, value, -time) %>%
+	mutate(
+		key=factor(key, levels=c("incidence", "mortality", "prevalence"),
+				   labels=c("measured upon infection", "measured upon recovery", "prevalence"))
+	)
+
+g1 <- ggplot(rr_data) +
+	geom_line(aes(time, value, col=key), lwd=1.5) +
+	scale_x_continuous("Time", expand=c(0, 0), limits=c(0, 20.5)) +
+	scale_y_continuous("Number of cases", limits=c(0, 1000)) +
+	theme(
+		legend.title = element_blank(),
+		legend.position = c(0.8, 0.9),
+		panel.grid = element_blank(),
+		panel.border = element_blank(),
+		axis.line = element_line()
+	)
+
+ggsave("example.pdf", g1, width=6, height=4)

### 2019-07-23 ms/ms.tex
very close

@@
 
-\title{A note on observational processes in epidemic models}
+\title{A note on observation processes in epidemic models}
 \author{Sang Woo Park and Benjamin M. Bolker}
@@
 
+\section*{Abstract}
+
 \section{Introduction}
@@
 transmission mechanism, estimate biologically relevant parameters, and 
-predict the course of an epidemic \citep{breto2009time}. In order to make a 
+predict the course of an outbreak \citep{breto2009time}. In order to make a 
 precise and accurate inference, disease modelers often focus on developing
@@
 modelers often neglect details of the observation process associated with disease
-case reports. Most disease models assume that new cases are reported
+case reports. Most disease models effectively assume that new cases are reported
 instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
-kennedy2018modeling, pons2018serotype}) or develops symptom; one exception is
-the model by \citep{he2009plug}, which assumes that infections are counted upon
-recovery (because diagnosed cases are put to bed rest and are effectively 
-no longer infectious).
+kennedy2018modeling, pons2018serotype}) or develops symptom (e.g., 
+\cite{bhadra2011malaria, king2015avoidable}); 
+a few models (e.g.,\cite{breto2009time, he2009plug, lin2016seasonality})
+assume that infections are counted upon recovery (because diagnosed cases
+are controlled and are effectively no longer infectious).
 
 Here, we use a simple Susceptible-Infected-Removed (SIR) model to study how 
-assumptions about the underlying observation process affect parameter estimates
+assumptions about the underlying observation processes affect parameter estimates
 of the SIR model. We show that making wrong assumptions about the timing of 
@@
 \end{equation}
-where $\beta$ is the contact rate, $\gamma$ is the recovery rate, and $N = S + I + R$ is
-the total population size. We define incidence at time $t$ as the number of newly infected
+where $\beta$ is the contact rate per unit time, $\gamma$ is the recovery rate per unit time, 
+and $N = S + I + R$ is the total population size. 
+We define incidence at time $t$ as the number of newly infected
 individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
@@
 Finally, we model under-reporting using a negative binomial distribution with
-mean $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
-an over dispersion parameter $\theta$. For convenience, we will refer to these two
+means $\rho i_1(t)$ and $\rho i_2(t)$, where $\rho$ is the reporting rate, and
+an over-dispersion parameter $\theta$. For convenience, we will refer to these two
 negative binomial models as infection model and recovery model hereafter; 
@@
 In order to introduce the problem, we first compare the dynamics of observed 
-incidence, $i_1(t)$ and $i_2(t)$, (Figure 1). Lags in reporting time delay 
+incidence curves, $i_1(t)$ and $i_2(t)$, (Figure 1). A lag in reporting time delays
 the observed epidemic peak timing and reduce the size of the peak. However, 
@@
 
-Note that incidence measured at these two periods are different from prevalence,
+Note that incidence, regardless of when it is measured, is different from prevalence,
 $I(t)$, which is defined as the number of \emph{currently} infected individuals;
@@
 the reporting time step is equal to the disease generation time. While it is
-uncommon, some models do not make a clear distinction between the two.
+uncommon, some models do not make a clear distinction between the two (e.g., 
+\cite{hooker2010parameterizing}).
 
@@
 \textbf{A comparison of incidence measured at two different time points in infection.}
-We used the following parameters to simulate the SIR model and generate the figure: 
+A deterministic SIR model is simulated using the following parameters: 
 $\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
@@
 \textbf{A comparison of incidence measured at two different time points in infection.}
-We used the following parameters to simulate the SIR model: 
+We simulate infection time series and recovery time series 100 times using the 
+following parameters:  
 $\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
-$\theta = 10$, and $\Delta t = 0.1$ time units.
+$\rho = 0.5$, $\theta = 10$, and $\Delta t = 0.1$ time units.
+For each simulation, we fit the infection model and the recovery model by
+(A) allowing for the mean infectious period to be estimated and (B) assuming
+that the mean infectious period is known.
+Coverage probabilities represent the proportion of confidence intervals
+that contain the true value of the basic reproductive number ($\mathcal R_0$).
 }
@@
 Disease modelers often assume that the mean infectious period of a disease
-is known and focus on estimating the basic reproduction number (or the 
-transmission rate). When we assume that the mean infectious period is known
+is known and focus on estimating the basic reproduction number (e.g.,
+\cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
+When we assume that the mean infectious period is known
 and try to estimate the remaining 4 parameters of the SIR model, fitting the
-wrong model results in clearer bias and lower coverage (Figure 2B); however, 
+wrong model results in clearer bias and a much lower coverage (Figure 2B); however, 
 bias from fitting the wrong model changes direction (e.g., fitting the infection
@@
 exponential growth rate ($r = \beta - \gamma$) and the mean infectious period ($1/\gamma$).
-When we fit the recovery model to infection time series, the model overestimates
+When we fit the recovery model to an infection time series, the model overestimates
 the initial growth rate in order to match the higher epidemic peak of the 
@@
 translates to higher basic reproduction number as we see in Figure 2B. When we allow
-the mean infectious period to vary, the model underestimates the mean infectious
-period (high $\gamma$), which also decreases the estimate of the basic reproduction number.
-Likewise, fitting the infection model to recovery time series overestimates the
+the mean infectious period to vary, the model also underestimates the mean infectious
+period (high $\gamma$), which decreases the overall estimate of the basic reproduction number.
+Likewise, fitting the infection model to a recovery time series overestimates the
 mean infectious period (low $\gamma$), increasing the estimate of the basic reproduction number;
 when the mean infectious period is fixed, low growth rate estimates (to match the
-lower epidemic peak of the recovery time series) decrease the estimate of the
+lower epidemic peak of the recovery time series) decrease the overall estimate of the
 basic reproduction number.
@@
 individuals become infected or when they recover. Although neither of these
-assumptions are realistic, they are broadly consistent with the assumptions
-of other state-of-art epidemic models. More realistic models may distinguish
-reported and unreported cases by adding a new state variable or by modeling
-an explicit delay kernel in reporting time.
+assumptions is realistic, they are broadly consistent with the assumptions
+of other state-of-art epidemic models (see Introduction). 
+More realistic models may distinguish reported and unreported
+(identified and unidentified) cases by adding
+new state variables \citep{browne2015modeling,webb2015model} 
+or by modeling an explicit delay kernel in reporting time 
+\citep{harris1990reporting, ferguson2001foot, ster2009epidemiological, 
+birrell2011bayesian, funk2018real}.
 
@@
 tried to infer underlying transmission mechanisms from historical mortality 
-data but assumed that deaths are recorded as soon as individuals die. This assumption
-does not account for the possibility that delays in reporting time of deaths can
+data but assumed that deaths are recorded as soon as individuals die 
+\citep{he2013inferring, didelot2017model, dean2018human}. 
+This assumption does not account for the possibility 
+that a delay in reporting time of deaths can
 change the shape of an epidemic curve: longer delays are expected to decrease
 the size of an epidemic peak and delay the timing of the peak (Figure 1). 
-These assumptions may have important effects on their conclusions about the underlying 
-transmission mechanism.
+These assumptions may affect their conclusions about the underlying 
+transmission mechanisms.
 

### 2019-07-23 ms/ms.tex
.

@@
 
+Epidemic models provide valuable tools for understanding how a disease
+spreads in a population. Many epidemic models focus on characterizing
+the underlying transmission mechanism but make naive assumptions about
+how infections are reported. In this note, we use a simple deterministic
+Susceptible-Infected-Removed (SIR) model to compare two popular assumptions 
+about disease incidence reports: individuals can report their infection as soon as
+they become infected or as soon as they recover. We show that making
+wrong assumptions about the underlying observation processes can result
+in biased estimates of the basic reproduction number and narrow confidence
+intervals.
+
 \section{Introduction}
@@
 predict the course of an outbreak \citep{breto2009time}. In order to make a 
-precise and accurate inference, disease modelers often focus on developing
+precise and accurate inference, disease modelers try to build
 more realistic process models. For example, the same London measles time 
@@
 
-The Susceptible-Infected-Removed (SIR) model describes how disease spreads in a
+The Susceptible-Infected-Removed (SIR) model describes how a disease spreads in a
 homogeneous population:
@@
 their infection when they become infected or when they recover. The observed incidence 
-measured upon infection ($i_1(t)$) can be defined by the integral:
+measured upon infection, $i_1(t)$, can be defined by the integral:
 \begin{equation}
-i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt,
+i_1(t) = \int_{t - \Delta t}^{t} \beta S \frac{I}{N} dt.
 \end{equation}
-where $\Delta t$ is the reporting time step.
 Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
@@
 the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
-incidence measured upon recovery ($i_2(t)$) can be defined by the integral:
+incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
 \begin{equation}
@@
 $i_2(t) = R(t) - R(t - \Delta t)$.
-Finally, we model under-reporting using a negative binomial distribution with
-means $\rho i_1(t)$ and $\rho i_2(t)$, where $\rho$ is the reporting rate, and
+Finally, we model under-reporting using a negative binomial distribution with a
+mean either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
 an over-dispersion parameter $\theta$. For convenience, we will refer to these two
 negative binomial models as infection model and recovery model hereafter; 
-likewise, we will refer to epidemic time series generated from these models 
+likewise, we will refer to epidemic time series generated from these
+negative binomial models 
 as infection time series and recovery time series.
@@
 
-In order to introduce the problem, we first compare the dynamics of observed 
-incidence curves, $i_1(t)$ and $i_2(t)$, (Figure 1). A lag in reporting time delays
+First, we compare the dynamics of incidence curves, $i_1(t)$ and $i_2(t)$, 
+(Figure 1). A lag in reporting time delays
 the observed epidemic peak timing and reduce the size of the peak. However, 
@@
 
-Note that incidence, regardless of when it is measured, is different from prevalence,
+Note that incidence is different from prevalence,
 $I(t)$, which is defined as the number of \emph{currently} infected individuals;
@@
 and try to estimate the remaining 4 parameters of the SIR model, fitting the
-wrong model results in clearer bias and a much lower coverage (Figure 2B); however, 
+wrong model results in a clearer bias and a much lower coverage (Figure 2B); however, 
 bias from fitting the wrong model changes direction (e.g., fitting the infection
@@
 Mathematical modeling of infectious disease outbreaks helps us understand 
-how disease spreads in a population. Disease modelers often put a lot of
-effort into capturing the transmission mechanism but make naive assumptions
-about how cases are reported. Here, we used a simple deterministic SIR model to show that 
+how disease spreads in a population; however, many epidemic models make 
+naive assumptions about how cases are reported. 
+Here, we used a deterministic SIR model to show that 
 making wrong assumptions about observational processes in epidemic models
@@
 tried to infer underlying transmission mechanisms from historical mortality 
-data but assumed that deaths are recorded as soon as individuals die 
+data but assumed that individual deaths are recorded as soon as individuals die 
 \citep{he2013inferring, didelot2017model, dean2018human}. 
 This assumption does not account for the possibility 
-that a delay in reporting time of deaths can
+that a delay in reporting of deaths can
 change the shape of an epidemic curve: longer delays are expected to decrease
 the size of an epidemic peak and delay the timing of the peak (Figure 1). 
-These assumptions may affect their conclusions about the underlying 
+Ignoring potential delays in reporting may affect their conclusions about the underlying 
 transmission mechanisms.

### 2019-07-26 ms/ms.tex
minor tweaks

@@
 \documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+\usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
 
@@
 
-First, we compare the dynamics of incidence curves, $i_1(t)$ and $i_2(t)$, 
+First, we compare the deterministic dynamics of incidence curves, 
+$i_1(t)$ and $i_2(t)$, 
 (Figure 1). A lag in reporting time delays
-the observed epidemic peak timing and reduce the size of the peak. However, 
+the observed epidemic peak timing and reduces the size of the peak. However, 
 these differences in the reporting time have small effect to the overall shape 
 of the epidemic curve. In the presence of observation and process error, we 
-would not expect to be able to distinguish between the two reporting processes.
+might not expect to be able to distinguish between the two reporting processes
+based on the time series alone.
 
-Note that incidence is different from prevalence,
-$I(t)$, which is defined as the number of \emph{currently} infected individuals;
-the dynamics of observed incidence depend on the reporting time step (because the sum of 
-true incidence is equal to the final size of an epidemic), whereas those of
-prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
-the reporting time step is equal to the disease generation time. While it is
-uncommon, some models do not make a clear distinction between the two (e.g., 
-\cite{hooker2010parameterizing}).
-
-\begin{figure}
+\begin{figure}[!t]
 \includegraphics[width=\textwidth]{../figure/example.pdf}
@@
 \end{figure}
+
+Note that incidence is different from prevalence,
+$I(t)$, which is defined as the number of \emph{currently} infected individuals;
+the observed dynamics of incidence depend on the reporting time step (because the sum of 
+true incidence is equal to the final size of an epidemic), whereas those of
+prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
+the reporting time step is equal to the disease generation time. While it is
+uncommon, some models do not make a clear distinction between the two (e.g., 
+\cite{hooker2010parameterizing}).
 
@@
 exponential growth rate ($r = \beta - \gamma$) and the mean infectious period ($1/\gamma$).
+In general, we expect delays in observation processes to make the observed epidemic 
+time series last longer and have smaller peaks (Figure 1).
 When we fit the recovery model to an infection time series, the model overestimates
-the initial growth rate in order to match the higher epidemic peak of the 
+the initial growth rate in order to match the bigger (and faster) epidemic peak of the 
 infection time series. When the mean infectious period is fixed, higher growth rate
 translates to higher basic reproduction number as we see in Figure 2B. When we allow
-the mean infectious period to vary, the model also underestimates the mean infectious
-period (high $\gamma$), which decreases the overall estimate of the basic reproduction number.
-Likewise, fitting the infection model to a recovery time series overestimates the
-mean infectious period (low $\gamma$), increasing the estimate of the basic reproduction number;
-when the mean infectious period is fixed, low growth rate estimates (to match the
-lower epidemic peak of the recovery time series) decrease the overall estimate of the
-basic reproduction number.
+the mean infectious period to vary, the model still over estimates the growth rate but
+also underestimates the mean infectious period (high $\gamma$), 
+which decreases the overall estimate of the basic reproduction number.
+Likewise, fitting the infection model to a recovery time series underestimates
+the growth rate to match the smaller (and slower) epidemic peak; this 
+underestimates the basic reproduction number when the mean infectious period is fixed. 
+When we allow the mean infectious period to vary, we overestimate the mean infectious
+period, which in turn increaseses the estimate of the basic reproduction number.
 

### 2019-07-30 ms/ms.tex
tweaks to writing

@@
 prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
-the reporting time step is equal to the disease generation time. While it is
-uncommon, some models do not make a clear distinction between the two (e.g., 
-\cite{hooker2010parameterizing}).
+the reporting time step is equal to the disease generation time. While they are relatively
+uncommon, some models do not make a clear distinction between a discrete-time observed 
+variable and a continuous-time latent variable (Table 1).
+
+\begin{table}
+\centering
+\begin{tabular}{l | c | c}
+\hline
+\textbf{Study} & \textbf{Disease} &  \textbf{Observed variables}\\
+\hline
+\cite{capistran2009parameter} & Respiratory syncytial virus & Number of hospital cases\\
+\cite{hooker2010parameterizing} & Measles & Number of reported cases\\
+\cite{eisenberg2013identifiability} & Cholera & Number of reported cases\\
+\cite{yang2013stability} & Hand-foot-mouth disease & Number of reported cases\\
+\cite{gonzalez2014fractional} & Influenza & Number of confirmed cases\\
+\hline
+\end{tabular}
+\caption{
+\textbf{List of studies that fit prevalence curves to incidence data.}
+Incidence, defined as the number of new cases between two reporting periods, depends
+on the reporting time step.
+Prevalence, defined as the number of currently infected individuals, does not depend
+on the reporting time step.
+Some models do not make a clear distinction between these two quantities.
+}
+\end{table}
 
@@
 More realistic models may distinguish reported and unreported
-(identified and unidentified) cases by adding
+(i.e., identified and unidentified) cases by adding
 new state variables \citep{browne2015modeling,webb2015model} 
@@
 
-While we have only considered observation processes associated with incidence 
+While we only considered observation processes associated with incidence 
 reporting in this study, it is equally as important to consider observation
@@
 data but assumed that individual deaths are recorded as soon as individuals die 
-\citep{he2013inferring, didelot2017model, dean2018human}. 
+\citep{he2013inferring, didelot2017model, dean2018human}; this includes
+\cite{kermack1927contribution} who approximated the reported \emph{number} of
+deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
 This assumption does not account for the possibility 
@@
 the size of an epidemic peak and delay the timing of the peak (Figure 1). 
-Ignoring potential delays in reporting may affect their conclusions about the underlying 
+Ignoring delays in reporting may affect their conclusions about the underlying 
 transmission mechanisms.
 
-Our study shows that seemingly irrelevant changes in the assumptions of an epidemic
-model affects the inference of infectious disease transmission.
+Our study shows that a seemingly negligible change in the assumptions of an epidemic
+model can affect the inference of infectious disease transmission.
 We caution disease modelers to be more mindful about their decisions in developing

### 2019-07-30 ms/ms.tex
keep tweaking

@@
 how infections are reported. In this note, we use a simple deterministic
-Susceptible-Infected-Removed (SIR) model to compare two popular assumptions 
+Susceptible-Infected-Removed (SIR) model to compare two common assumptions 
 about disease incidence reports: individuals can report their infection as soon as
@@
 predict the course of an outbreak \citep{breto2009time}. In order to make a 
-precise and accurate inference, disease modelers try to build
+precise and accurate inference, disease modelers have sought to build
 more realistic process models. For example, the same London measles time 
-series from the prevaccination era has been analyzed using multiple models -- 
+series from the prevaccination era has been analyzed many times -- 
 these models account for time-varying transmission rates \citep{fine1982measles}, 
@@
 modelers often neglect details of the observation process associated with disease
-case reports. Most disease models effectively assume that new cases are reported
+case reports. Many disease models effectively assume that new cases are reported
 instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
@@
 \cite{bhadra2011malaria, king2015avoidable}); 
-a few models (e.g.,\cite{breto2009time, he2009plug, lin2016seasonality})
+some models (e.g., \cite{breto2009time, he2009plug, lin2016seasonality})
 assume that infections are counted upon recovery (because diagnosed cases
@@
 likewise, we will refer to epidemic time series generated from these
-negative binomial models 
+negative binomial models with two different means ($\rho i_1(t)$ and $\rho i_2(t)$)
 as infection time series and recovery time series.
@@
 First, we compare the deterministic dynamics of incidence curves,
-$i_1(t)$ and $i_2(t)$, 
-(Figure 1). A lag in reporting time delays
+$i_1(t)$ and $i_2(t)$, in Figure 1. A lag in reporting time delays
 the observed epidemic peak timing and reduces the size of the peak. However, 
@@
 In order to understand how assumptions about the timing of case reporting affect 
-parameter estimates of the SIR model, we simulate observed epidemic curves 
-(i.e., infection time series and recovery time series), measured upon both infection and
-recovery, 100 times and fit both models (i.e., infection model and recovery model) to each 
+parameter estimates of the SIR model, we simulate observed epidemic time series
+(infection time series and recovery time series) 
+100 times and fit both infection and recovery models to each 
 time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
@@
 \cite{hooker2010parameterizing, lin2016seasonality, pons2018serotype}). 
-When we assume that the mean infectious period is known
+When we assume that the true value of the mean infectious period is known
 and try to estimate the remaining 4 parameters of the SIR model, fitting the
 wrong model results in a clearer bias and a much lower coverage (Figure 2B); however, 
-bias from fitting the wrong model changes direction (e.g., fitting the infection
+bias changes direction (e.g., fitting the infection
 model to recovery model overestimates the reproduction number instead).
@@
 that a delay in reporting of deaths can
-change the shape of an epidemic curve: longer delays are expected to decrease
-the size of an epidemic peak and delay the timing of the peak (Figure 1). 
-Ignoring delays in reporting may affect their conclusions about the underlying 
+change the shape of an epidemic curve: delayed reporting is expected to decrease
+the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1). 
+Ignoring delays in reporting may affect conclusions about the underlying 
 transmission mechanisms.

### 2019-07-30 ms/ms.tex
more tweaks

@@
 
-We assumed that newly infected cases are reported instantaneously when
-individuals become infected or when they recover. Although neither of these
-assumptions is realistic, they are broadly consistent with the assumptions
-of other state-of-art epidemic models (see Introduction). 
+We compared two scenarios in which newly infected cases are reported 
+instantaneously (1) when individuals become infected or
+(2) when they recover. Although neither of these
+assumptions is realistic, many epidemic models still rely on these
+assumptions (see Introduction). 
 More realistic models may distinguish reported and unreported
@@
 
-While we only considered observation processes associated with incidence 
-reporting in this study, it is equally as important to consider observation
-processes in mortality reporting. In particular, many disease modelers have
+We considered observation processes associated with incidence 
+reporting; however, we expect observation processes to be just as important
+in mortality reporting. In particular, many disease modelers have
 tried to infer underlying transmission mechanisms from historical mortality 
@@
 deaths per week from plague with instantaneous death \emph{rates} ($dR/dt$).
-This assumption does not account for the possibility 
+These frameworks do not account for the possibility 
 that a delay in reporting of deaths can
-change the shape of an epidemic curve: delayed reporting is expected to decrease
+change the shape of an epidemic curve: delays in case reports can decrease
 the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1).
-Ignoring delays in reporting may affect conclusions about the underlying 
+
+There is some empirical evidence that delays in case notification alters the
+shape of an epidemic curve and affects our understanding of disease transmission:
+(1) Postal delays in measles notification during holidays lead to sudden
+epidemic peaks after holidays \citep{fine1982measles}; 
+(2) Delays in reports of the foot-and-mouth infection affects the effectiveness
+of slaughter policies on epidemic intervention;
+and (3) Date of wills written and date of wills probated during
+the plague epidemics in London give completely different looking epidemic curves
+due to extremely long delays (mean of 2.9 years) between the two \citep{bushby2019wills}.
+Ignoring such delays in reporting may affect conclusions about the underlying 
 transmission mechanisms.

### 2019-07-31 ms/ms.tex
minor tweaks

@@
-\documentclass{article}\usepackage[]{graphicx}\usepackage[]{color}
+\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
 \usepackage[top=1in,left=1in, right = 1in, footskip=1in]{geometry}
@@
 
-First, we compare the deterministic dynamics of incidence curves,
-$i_1(t)$ and $i_2(t)$, in Figure 1. A lag in reporting time delays
-the observed epidemic peak timing and reduces the size of the peak. However, 
-these differences in the reporting time have small effect to the overall shape 
-of the epidemic curve. In the presence of observation and process error, we 
-might not expect to be able to distinguish between the two reporting processes
-based on the time series alone.
-
 \begin{figure}[!t]
@@
 \end{figure}
+
+First, we compare the deterministic dynamics of incidence curves,
+$i_1(t)$ and $i_2(t)$, in Figure 1. A lag in reporting time delays
+the observed epidemic peak timing and reduces the size of the peak. However, 
+these differences in the reporting time have small effect to the overall shape 
+of the epidemic curve. In the presence of observation and process error, we 
+might not expect to be able to distinguish between the two reporting processes
+based on the time series alone.
 
@@
 When we allow the mean infectious period to vary, we overestimate the mean infectious
-period, which in turn increaseses the estimate of the basic reproduction number.
+period, which in turn increases the estimate of the basic reproduction number.
 

### 2019-07-31 ms/ms.tex
tweaks

@@
 new state variables \citep{browne2015modeling,webb2015model} 
-or by modeling an explicit delay kernel in reporting time 
-\citep{harris1990reporting, ferguson2001foot, ster2009epidemiological, 
-birrell2011bayesian, funk2018real}.
+or by modeling an explicit delay distribution in reporting time 
+\citep{harris1990reporting, ferguson2001foot, goldstein2009reconstructing,
+ster2009epidemiological, birrell2011bayesian, funk2018real}.
 
@@
 
+Here, we assumed that the underlying transmission process is deterministic; this 
+assumes that all error can be explained by observation errors alone. 
+We chose to study a deterministic model SIR for computational efficiency; 
+we do not recommend using deterministic models for real outbreak analyses.
+Ignoring process errors (i.e., stochasticity in the transmission process) can lead to
+overly confident results \citep{king2015avoidable}. Using stochastic models may
+even give better coverage probabilities when wrong observation models are used. 
+Nonetheless, misspecifying the observation model can still affect conclusions from 
+stochastic models and introduce bias.
+
 Our study shows that a seemingly negligible change in the assumptions of an epidemic

### 2019-08-15 ms/ms.tex
tweaks

@@
 Epidemic models provide valuable tools for understanding how a disease
-spreads in a population. Many epidemic models focus on characterizing
+spreads in a population. Many disease modelers focus on characterizing
 the underlying transmission mechanism but make naive assumptions about
@@
 they become infected or as soon as they recover. We show that making
-wrong assumptions about the underlying observation processes can result
-in biased estimates of the basic reproduction number and narrow confidence
-intervals.
+wrong assumptions about the underlying observation processes can introduce
+small bias to the estimates of the basic reproduction number and give
+narrow confidence intervals.
 
@@
 
-Despite the amount of effort put into developing \emph{better} process models, disease 
+Despite the amount of effort put into developing \emph{better} process models, 
+disease 
 modelers often neglect details of the observation process associated with disease
@@
 instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
-kennedy2018modeling, pons2018serotype}) or develops symptom (e.g., 
-\cite{bhadra2011malaria, king2015avoidable}); 
+kennedy2018modeling, pons2018serotype}) or when an individual becomes symptomatic
+(e.g., \cite{bhadra2011malaria, king2015avoidable}); 
 some models (e.g., \cite{breto2009time, he2009plug, lin2016seasonality})
@@
 of the epidemic curve. In the presence of observation and process error, we 
-might not expect to be able to distinguish between the two reporting processes
+\emph{might} not expect to be able to distinguish between the two reporting processes
 based on the time series alone.
@@
 the reporting time step is equal to the disease generation time. While they are relatively
-uncommon, some models do not make a clear distinction between a discrete-time observed 
-variable and a continuous-time latent variable (Table 1).
-
-\begin{table}
-\centering
-\begin{tabular}{l | c | c}
-\hline
-\textbf{Study} & \textbf{Disease} &  \textbf{Observed variables}\\
-\hline
-\cite{capistran2009parameter} & Respiratory syncytial virus & Number of hospital cases\\
-\cite{hooker2010parameterizing} & Measles & Number of reported cases\\
-\cite{eisenberg2013identifiability} & Cholera & Number of reported cases\\
-\cite{yang2013stability} & Hand-foot-mouth disease & Number of reported cases\\
-\cite{gonzalez2014fractional} & Influenza & Number of confirmed cases\\
-\hline
-\end{tabular}
-\caption{
-\textbf{List of studies that fit prevalence curves to incidence data.}
-Incidence, defined as the number of new cases between two reporting periods, depends
-on the reporting time step.
-Prevalence, defined as the number of currently infected individuals, does not depend
-on the reporting time step.
-Some models do not make a clear distinction between these two quantities.
-}
-\end{table}
+uncommon, some models do not make a clear distinction between prevalence and 
+incidence 
+\citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.
 
@@
 and try to estimate the remaining 4 parameters of the SIR model, fitting the
-wrong model results in a clearer bias and a much lower coverage (Figure 2B); however, 
-bias changes direction (e.g., fitting the infection
-model to recovery model overestimates the reproduction number instead).
+wrong model results in a clearer bias (in opposite directions) 
+and a much lower coverage (Figure 2B).
 
@@
 Mathematical modeling of infectious disease outbreaks helps us understand 
-how disease spreads in a population; however, many epidemic models make 
-naive assumptions about how cases are reported. 
+how disease spreads in a population; however, epidemic models often make 
+simple assumptions about how cases are reported. 
 Here, we used a deterministic SIR model to show that 
@@
 We considered observation processes associated with incidence 
-reporting; however, we expect observation processes to be just as important
-in mortality reporting. In particular, many disease modelers have
+reporting; however, we expect observation processes to be just as 
+(if not more) important in analyzing mortality data. In particular,
+many disease modelers have
 tried to infer underlying transmission mechanisms from historical mortality 
@@
 that a delay in reporting of deaths can
-change the shape of an epidemic curve: delays in case reports can decrease
+change the shape of an epidemic curve -- delays in case reports can decrease
 the size of the observed epidemic peak and delay the observed timing of the peak (Figure 1).
@@
 model can affect the inference of infectious disease transmission.
+While the amount of bias introduced from the misspecification of the observation
+model can be fairly small (e.g., less than 5\% in our examples), any modelers
+trying to make serious forecasts should be aware of potential biases.
 We caution disease modelers to be more mindful about their decisions in developing

### 2019-08-15 figure/compare_deterministic.R

@@ -47,11 +47,16 @@ for (i in 1:8) {
 summdata <- summlist %>%
 	bind_rows
 
+summdata %>%
+	group_by(data, fit, inf) %>%
+	summarize(bias=(mean(estimate)-2)/2)
+
 g1 <- ggplot(filter(summdata, inf=="estimated")) +
 	geom_boxplot(aes(x=data, y=estimate, fill=fit), width=0.5) +
 	geom_hline(yintercept=2, lty=2) +
 	scale_x_discrete("Simulated timing") +
-	scale_y_continuous("Basic reproductive number") +
+	scale_y_continuous("Basic reproductive number",
+					   limits=c(1.7, 2.6)) +
 	theme(
 		legend.position="none"
 	)
@@ -74,7 +79,7 @@ g2 <- ggplot(filter(coverdata, inf=="estimated")) +
 	)
 
 g3 <- g1 %+% filter(summdata, inf=="fixed")
-
+	
 g4 <- g2 %+% filter(coverdata, inf=="fixed") +
 	theme(legend.position="none")
 

### 2019-08-15 figure/compare_stochastic.R

@@ -1,88 +1,0 @@
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
-scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
-
-load("../data/stochastic.rda")
-
-allpred <- allres <- allsum <- allmle <- list()
-
-load("../analysis/fakedata_incidence.rda")
-
-allpred$incidence <- predres
-allres$incidence <- reslist
-allsum$incidence <- sumlist
-allmle$incidence <- mle
-
-load("../analysis/fakedata_mortality.rda")
-
-allpred$mortality <- predres
-allres$mortality <- reslist
-allsum$mortality <- sumlist
-allmle$mortality <- mle
-
-summdata <- allsum %>%
-	lapply(bind_rows) %>%
-	bind_rows(.id="type") %>%
-	mutate(
-		type=factor(type, levels=c("incidence", "mortality"), labels=c("infection", "recovery"))
-	) %>%
-	filter(
-		R0 < 2.5
-	)
-
-horizontal <- data.frame(
-	hh=c(max(predict(loess(logLik~R0, data=filter(summdata, type=="infection")))), 
-		 max(predict(loess(logLik~R0, data=filter(summdata, type=="recovery")))))-qchisq(0.95, 1)/2,
-	type=c("infection", "recovery")
-)
-
-
-
-g1 <- ggplot(summdata) +
-	geom_smooth(aes(R0, logLik), lty=1, se=FALSE, method="loess", span=0.5, col="black", lwd=0.5) +
-	geom_point(aes(R0, logLik), shape=1, size=2) +
-	geom_hline(data=horizontal, aes(yintercept=hh), lty=2) +
-	geom_vline(data=data.frame(type="infection"), aes(xintercept=1.34), lty=2) +
-	geom_vline(data=data.frame(type="infection"), aes(xintercept=2.20), lty=2) +
-	geom_vline(data=data.frame(type="recovery"), aes(xintercept=1.64), lty=2) +
-	geom_vline(data=data.frame(type="recovery"), aes(xintercept=2.25), lty=2) +
-	scale_x_continuous("Basic reproductive number") +
-	scale_y_continuous("Profile log-likelihood") +
[…]

### 2019-11-08 ms/ms.tex
acknowledge david
maybe add stochastic things

@@
 
-Epidemic models provide valuable tools for understanding how a disease
-spreads in a population. Many disease modelers focus on characterizing
-the underlying transmission mechanism but make naive assumptions about
+Many disease models focus on characterizing
+the underlying transmission mechanism but make simple, possibly naive assumptions about
 how infections are reported. In this note, we use a simple deterministic
@@
 about disease incidence reports: individuals can report their infection as soon as
-they become infected or as soon as they recover. We show that making
-wrong assumptions about the underlying observation processes can introduce
-small bias to the estimates of the basic reproduction number and give
+they become infected or as soon as they recover. We show that
+incorrect assumptions about the underlying observation processes can bias
+estimates of the basic reproduction number and lead to overly
 narrow confidence intervals.
@@
 
-Mechanistic analyses of epidemic time series allow us to infer the underlying 
-transmission mechanism, estimate biologically relevant parameters, and 
-predict the course of an outbreak \citep{breto2009time}. In order to make a 
-precise and accurate inference, disease modelers have sought to build
-more realistic process models. For example, the same London measles time 
-series from the prevaccination era has been analyzed many times -- 
-these models account for time-varying transmission rates \citep{fine1982measles}, 
+Mechanistic analyses of epidemic time series allow us to make inference about the underlying 
+transmission mechanisms, estimate biologically relevant parameters, and 
+predict the course of an outbreak \citep{breto2009time}. In order to make 
+precise and accurate inferences, disease modelers have sought to build
+more realistic process models. For example, the time series of reported
+measles cases from London in the prevaccination era has been analyzed many times, using
+variety of models accounting for time-varying transmission rates \citep{fine1982measles}, 
 realistic age structure \citep{schenzle1984age},
-metapopulation structure \citep{xia2004measles}, continuous-time infection process 
+metapopulation structure \citep{xia2004measles}, continuous-time infection processes 
 \citep{cauchemez2008likelihood}, and extra-demographic variability \citep{he2009plug}.
 
-Despite the amount of effort put into developing \emph{better} process models, 
-disease 
-modelers often neglect details of the observation process associated with disease
-case reports. Many disease models effectively assume that new cases are reported
+Despite the amount of effort put into developing better process models, 
+disease modelers often neglect details of the observation processes associated with
+new disease case reports (often referred to as incidence time series).
+Many disease models effectively assume that new cases are reported
 instantaneously when an individual is infected (e.g., \cite{martinez2016differential, 
@@
 are controlled and are effectively no longer infectious). 
+\bmb{Maybe mention King et al. -- completely ignored observation processes}
+
+Note that incidence (the number of \emph{newly} infected individuals) is different from 
+prevalence (the number of \emph{currently} infected individuals) \citep{bjornstad2018epidemics};
+the observed dynamics of incidence depend on the reporting time step (because the sum of 
+true incidence is equal to the final size of an epidemic), whereas those of
+prevalence do not. 
+While they are relatively
+uncommon, some models do not make a clear distinction between prevalence and 
+incidence 
+\citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.
 
@@
 of the SIR model. We show that making wrong assumptions about the timing of 
-incidence reports can lead to biases in parameter estimates and narrow 
+incidence reports can lead to biased parameter estimates and overly narrow 
 confidence intervals.
@@
 individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
-the reporting time step. We expect infected individuals to report their infection some time
-after their infection; the number of reported cases defines the \emph{observed} incidence.
+the reporting time step. We expect infected cases to be reported some time after infection;
+the number of reported cases during a time period defines the \emph{observed} incidence.
 
@@
 \end{equation}
-Alternatively, we can keep track of cumulative incidence, $C$, by adding a 
+Equivalently, we can keep track of cumulative incidence, $C$, by adding a 
 new state variable described by $dC/dt = \beta S I/N$ and taking the difference between 
@@
 $i_2(t) = R(t) - R(t - \Delta t)$.
-Finally, we model under-reporting using a negative binomial distribution with a
-mean either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
+Finally, we model reporting error using a negative binomial distribution with a
+mean of either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
 an over-dispersion parameter $\theta$. For convenience, we will refer to these two
 negative binomial models as infection model and recovery model hereafter; 
-likewise, we will refer to epidemic time series generated from these
+similarly, we will refer to epidemic time series generated from these
 negative binomial models with two different means ($\rho i_1(t)$ and $\rho i_2(t)$)
@@
 number $\mathcal R_0 = \beta/\gamma$, mean infectious period $1/\gamma$, 
-reporting rate $\rho$, an over dispersion parameter $\theta$ and the initial
-proportion of the infected individuals $i_0$. Initial proportion of
+reporting rate $\rho$, an overdispersion parameter $\theta$, and the initial
+proportion of the infected individuals $i_0$. The initial proportion of
 susceptible individuals is assumed to be $1 - i_0$. The total population
@@
 
-First, we compare the deterministic dynamics of incidence curves,
-$i_1(t)$ and $i_2(t)$, in Figure 1. A lag in reporting time delays
-the observed epidemic peak timing and reduces the size of the peak. However, 
-these differences in the reporting time have small effect to the overall shape 
+Figure 1 compares the deterministic dynamics of incidence curves,
+$i_1(t)$ and $i_2(t)$, and a prevalence curve $I(t)$. A lag in reporting time delays
+the observed epidemic peak timing and slightly reduces the size of the peak. However, 
+these differences in the reporting time have little effect on the overall shape 
 of the epidemic curve. In the presence of observation and process error, we
-\emph{might} not expect to be able to distinguish between the two reporting processes
-based on the time series alone.
-
-Note that incidence is different from prevalence,
-$I(t)$, which is defined as the number of \emph{currently} infected individuals;
-the observed dynamics of incidence depend on the reporting time step (because the sum of 
-true incidence is equal to the final size of an epidemic), whereas those of
-prevalence do not. We expect the dynamics of incidence and prevalence to be similar only when
-the reporting time step is equal to the disease generation time. While they are relatively
-uncommon, some models do not make a clear distinction between prevalence and 
-incidence 
-\citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.
+do not expect to be able to distinguish between the two reporting processes
+based on the time series alone (and we rarely know \emph{a priori} the delay distribution between when an individual is infected and when that case is reported).
+Therefore, one might naively expect assumptions about the timing of case reporting to have negligible effect on inference.
 
@@
 time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
-and its coverage probability, defined as the proportion of confidence intervals that
+and the coverage of our confidence intervals, defined as the proportion of confidence intervals that
 contain the true value (95\% confidence interval is expected to contain the true value
-95\% of the time by definition). We summarize the results in Figure 2.
+95\% of the time by definition). Figure 2 summarizes the results.
 
@@
 For each simulation, we fit the infection model and the recovery model by
-(A) allowing for the mean infectious period to be estimated and (B) assuming
+(A) estimating the mean infectious period and (B) assuming
 that the mean infectious period is known.
@@
 and a much lower coverage (Figure 2B).
+\swp{Cite elderd}
 
@@
 infection time series. When the mean infectious period is fixed, higher growth rate
-translates to higher basic reproduction number as we see in Figure 2B. When we allow
-the mean infectious period to vary, the model still over estimates the growth rate but
+translates to higher basic reproduction number, as Figure 2B shows. When we allow
+the mean infectious period to vary, the model still overestimates the growth rate but
 also underestimates the mean infectious period (high $\gamma$), 
 which decreases the overall estimate of the basic reproduction number.
-Likewise, fitting the infection model to a recovery time series underestimates
+Similarly, fitting the infection model to a recovery time series underestimates
 the growth rate to match the smaller (and slower) epidemic peak; this 
@@
 
+\bmb{Not including process error. Discuss?}
+\bmb{Constant reporting delays don't matter, if system is autonomous}
+
 Mathematical modeling of infectious disease outbreaks helps us understand 
@@
 Here, we used a deterministic SIR model to show that 
-making wrong assumptions about observational processes in epidemic models
-can give biased estimates of the basic reproduction number and narrow
+incorrect assumptions about observational processes in epidemic models
+can bias estimates of the basic reproduction number and lead to overly narrow
 confidence intervals.
+\bmb{repetitive}
 
@@
 reporting; however, we expect observation processes to be just as 
-(if not more) important in analyzing mortality data. In particular,
-many disease modelers have
+(if not more) important in analyzing mortality data. Many disease modelers have
 tried to infer underlying transmission mechanisms from historical mortality 
 data but assumed that individual deaths are recorded as soon as individuals die 
-\citep{he2013inferring, didelot2017model, dean2018human}; this includes
+\citep{he2013inferring, didelot2017model, dean2018human}; this includes the classic work by
 \cite{kermack1927contribution} who approximated the reported \emph{number} of
@@
 trying to make serious forecasts should be aware of potential biases.
-We caution disease modelers to be more mindful about their decisions in developing
-epidemic models and the implications of their model assumptions.
+We caution disease modelers to carefully consider the implications of their model assumptions when developing epidemic models.
+
+\section*{Acknowledgements}
+
+We thank David Earn for providing useful comments.
 

### 2019-11-08 figure/compare_stochastic.R

@@ -1,0 +1,74 @@
+library(dplyr)
+library(ggplot2); theme_set(theme_bw())
+library(gridExtra)
+
+scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
+scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+load("../data/stochastic.rda")
+
+allpred <- allres <- allsum <- allmle <- list()
+
+load("../analysis/fakedata_incidence.rda")
+
+allpred$incidence <- predres
+allres$incidence <- reslist
+allsum$incidence <- sumlist
+allmle$incidence <- mle
+
+load("../analysis/fakedata_mortality.rda")
+
+allpred$mortality <- predres
+allres$mortality <- reslist
+allsum$mortality <- sumlist
+allmle$mortality <- mle
+
+summdata <- allsum %>%
+  lapply(bind_rows) %>%
+  bind_rows(.id="type") %>%
+  mutate(
+    type=factor(type, levels=c("incidence", "mortality"), 
+                labels=c("measured upon infection", "measured upon recovery"))
+  ) %>%
+  filter(
+    R0 < 2.5
+  )
+
+horizontal <- data.frame(
+  hh=c(max(predict(loess(logLik~R0, data=filter(summdata, type=="measured upon infection")))), 
+       max(predict(loess(logLik~R0, data=filter(summdata, type=="measured upon recovery")))))-qchisq(0.95, 1)/2,
+  type=c("measured upon infection", "measured upon recovery")
+)
+
+mledata <- allmle %>% 
+  lapply(function(x){
+    data.frame(
+      R0=coef(x)[1]
+    )
+  }) %>%
+  bind_rows(.id="type") %>%
+  mutate(
+    type=factor(type, levels=c("incidence", "mortality"), 
+                labels=c("measured upon infection", "measured upon recovery"))
+  )
+
+g1 <- ggplot(summdata) +
+  geom_smooth(aes(R0, logLik), lty=1, se=FALSE, method="loess", span=0.5, col="black", lwd=0.5) +
+  geom_point(aes(R0, logLik), shape=1, size=2) +
+  geom_hline(data=horizontal, aes(yintercept=hh), lty=2) +
+  geom_vline(data=mledata, aes(xintercept=R0)) +
+  geom_rect(data=data.frame(type="measured upon infection"),
[…]

### 2019-11-08 figure/example.R

@@ -2,13 +2,14 @@ library(ggplot2); theme_set(theme_bw())
 library(tidyr)
 library(dplyr)
 library(deSolve)
+library(directlabels)
 source("../R/model.R")
 source("../R/run.R")
 
 scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
 
-rr <- run_sir()
+rr <- run_sir(tmax=22)
 
 rr_data <- rr %>%
 	gather(key, value, -time) %>%
@@ -18,12 +19,14 @@ rr_data <- rr %>%
 	)
 
 g1 <- ggplot(rr_data) +
-	geom_line(aes(time, value, col=key), lwd=1.5) +
-	scale_x_continuous("Time", expand=c(0, 0), limits=c(0, 20.5)) +
-	scale_y_continuous("Number of cases", limits=c(0, 1000)) +
+	geom_line(aes(time, value, col=key, lty=key), lwd=1.5) +
+  scale_x_continuous("Time", expand=c(0, 0), limits=c(0, 22)) +
+	scale_y_log10("Number of cases", limits=c(0.5, 25000)) +
+  scale_linetype_manual(values=c(1, 5, 4)) +
 	theme(
 		legend.title = element_blank(),
-		legend.position = c(0.8, 0.9),
+		legend.position = c(0.78, 0.9),
+		legend.key.width = unit(4, "line"),
 		panel.grid = element_blank(),
 		panel.border = element_blank(),
 		axis.line = element_line()

### 2019-11-25 ms/ms.tex
edits

@@
 Note that incidence (the number of \emph{newly} infected individuals) is different from 
-prevalence (the number of \emph{currently} infected individuals) \citep{bjornstad2018epidemics};
+prevalence (the number of \emph{currently} infected individuals) \citep{bjornstad2018epidemics}:
 the observed dynamics of incidence depend on the reporting time step (because the sum of 
@@
 assumptions about the underlying observation processes affect parameter estimates
-of the SIR model. We show that making wrong assumptions about the timing of 
+of the SIR model. We show that making incorrect assumptions about the timing of 
 incidence reports can lead to biased parameter estimates and overly narrow 
@@
 and $N = S + I + R$ is the total population size. 
-We define incidence at time $t$ as the number of newly infected
+We define \emph{true} incidence at time $t$ as the number of newly infected
 individuals that are infected between time $t- \Delta t$ and time $t$, where $\Delta t$ is
 the reporting time step. We expect infected cases to be reported some time after infection;
-the number of reported cases during a time period defines the \emph{observed} incidence.
+the number of reported cases during a time period defines the \emph{observed} incidence
 
@@
 the two consecutive reporting periods: $i_1(t) = C(t) - C(t-\Delta t)$. Likewise, 
-incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
+the observed incidence measured upon recovery, $i_2(t)$, can be defined by the integral:
 \begin{equation}
@@
 $i_2(t) = R(t) - R(t - \Delta t)$.
-Finally, we model reporting error using a negative binomial distribution with a
+Finally, we model observation error using a negative binomial distribution with a
 mean of either $\rho i_1(t)$ or $\rho i_2(t)$, where $\rho$ is the reporting rate, and
@@
 number $\mathcal R_0 = \beta/\gamma$, mean infectious period $1/\gamma$, 
-reporting rate $\rho$, an overdispersion parameter $\theta$, and the initial
+reporting rate $\rho$, the overdispersion parameter $\theta$, and the initial
 proportion of the infected individuals $i_0$. The initial proportion of
@@
 \textbf{A comparison of incidence measured at two different time points in infection.}
-A deterministic SIR model is simulated using the following parameters: 
+(A) 
+A deterministic simulation of the SIR model using the following parameters: 
 $\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
 and $\Delta t = 0.1$ time units.
+(B-C) Effects of $\mathcal R_0$ on the relative difference in the size of the epidemic peak ($1 - [\max i_2(t)]/[\max i_1(t)]$) and the difference in the peak timing ($\hat{t}_2 - \hat{t}_1$ where $i_k(\hat{t}_k) = \max i_k(t)$ for $k = 1, 2$) due to lags in reporting time.
+Black circles: simulation results of the deterministic SIR model.
+Black lines: locally estimated scatterplot smoothing (LOESS) curves fitted to simulation results.
+Remaining parameters are held constant ($1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$, and $\Delta t = 0.1$ time units) throughout simulations.
 }
@@
 
-Figure 1 compares the deterministic dynamics of incidence curves,
-$i_1(t)$ and $i_2(t)$, and a prevalence curve $I(t)$. A lag in reporting time delays
-the observed epidemic peak timing and slightly reduces the size of the peak. However, 
+Figure 1A compares the deterministic dynamics of two incidence curves,
+$i_1(t)$ and $i_2(t)$, and a prevalence curve $I(t)$ for $\mathcal R_0 = 2$. 
+A lag in reporting time delays
+the timing of the observed epidemic peak and reduces the size of that peak.
+As $\mathcal R_0$ increases, the difference in the size of the peaks increases (Figure 1B)
+but the difference in the timing of the peaks decreases (Figure 1C).
+For example, when $\mathcal R_0 = 5$, such delay in the reporting of new cases can reduce the size of the observed epidemic peak by almost 50\%.
+
+For small values of $\mathcal R_0$, 
 these differences in the reporting time have little effect on the overall shape 
@@
 (infection time series and recovery time series) 
-100 times and fit both infection and recovery models to each 
+100 times with $\mathcal R_0 = 2$ and fit both infection and recovery models to each 
 time series. We compare the estimates of the basic reproductive number $\mathcal R_0$,
@@
 When we assume that the true value of the mean infectious period is known
-and try to estimate the remaining 4 parameters of the SIR model, fitting the
-wrong model results in a clearer bias (in opposite directions) 
-and a much lower coverage (Figure 2B).
-\swp{Cite elderd}
+and try to estimate the remaining 4 parameters of the SIR model (Figure 2B), fitting the
+incorrect model results in a clearer bias (in opposite directions)
+and a much lower coverage (cf. \cite{elderd2006uncertainty}).
 
@@
 simple assumptions about how cases are reported. 
-Here, we used a deterministic SIR model to show that 
-incorrect assumptions about observational processes in epidemic models
-can bias estimates of the basic reproduction number and lead to overly narrow
-confidence intervals.
-\bmb{repetitive}
+We used a deterministic SIR model to show that
+assumptions about observation processes can
+affect the estimates of the basic reproduction number and their associated confidence intervals.
 
@@
 
-There is some empirical evidence that delays in case notification alters the
-shape of an epidemic curve and affects our understanding of disease transmission:
-(1) Postal delays in measles notification during holidays lead to sudden
-epidemic peaks after holidays \citep{fine1982measles}; 
-(2) Delays in reports of the foot-and-mouth infection affects the effectiveness
-of slaughter policies on epidemic intervention;
-and (3) Date of wills written and date of wills probated during
-the plague epidemics in London give completely different looking epidemic curves
-due to extremely long delays (mean of 2.9 years) between the two \citep{bushby2019wills}.
-Ignoring such delays in reporting may affect conclusions about the underlying 
-transmission mechanisms.
-
 Here, we assumed that the underlying transmission process is deterministic; this 
@@
 overly confident results \citep{king2015avoidable}. Using stochastic models may
-even give better coverage probabilities when wrong observation models are used. 
+give better coverage probabilities even  when wrong observation models are used.
 Nonetheless, misspecifying the observation model can still affect conclusions from 

### 2019-11-25 figure/example.R

@@ -1,6 +1,7 @@
 library(ggplot2); theme_set(theme_bw())
 library(tidyr)
 library(dplyr)
+library(gridExtra)
 library(deSolve)
 library(directlabels)
 source("../R/model.R")
@@ -8,6 +9,30 @@ source("../R/run.R")
 
 scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
 scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
+
+R0vec <- seq(1.1, 5, by=0.1)
+fixed_inf_list <- vector('list', length(R0vec))
+
+for (i in 1:length(R0vec)) {
+  print(i)
+  newpar <- c(
+    R0=R0vec[i],
+    gamma=1,
+    N=1e5,
+    rho=0.5
+  )
+  
+  rr <- run_sir(param=newpar, tmax=200, tlength=0.1)
+  
+  fixed_inf_list[[i]] <- data.frame(
+    peakdiff=1-max(rr$mortality)/max(rr$incidence),
+    timediff=(which.max(rr$mortality)-which.max(rr$incidence))*0.1,
+    R0=R0vec[i]
+  )
+}
+
+fixed_inf_data <- fixed_inf_list %>%
+  bind_rows
 
 rr <- run_sir(tmax=22)
 
@@ -23,6 +48,7 @@ g1 <- ggplot(rr_data) +
   scale_x_continuous("Time", expand=c(0, 0), limits=c(0, 22)) +
 	scale_y_log10("Number of cases", limits=c(0.5, 25000)) +
   scale_linetype_manual(values=c(1, 5, 4)) +
+  ggtitle("A") +
 	theme(
 		legend.title = element_blank(),
 		legend.position = c(0.78, 0.9),
@@ -32,4 +58,31 @@ g1 <- ggplot(rr_data) +
 		axis.line = element_line()
 	)
 
-ggsave("example.pdf", g1, width=6, height=4)
+g2 <- ggplot(fixed_inf_data) +
+  geom_smooth(aes(R0, peakdiff), lwd=1, se=FALSE, col=1) +
+  geom_point(aes(R0, peakdiff), pch=1, size=3) +
+  ggtitle("B") +
+  scale_x_continuous("Basic reproduction number", limits=c(1, 5.1), expand=c(0, 0)) +
+  scale_y_continuous("Relative difference in the epidemic peak", limits=c(0, 0.5), expand=c(0, 0)) +
+  theme(
[…]

### 2019-11-27 ms/ms.tex
minor updates

@@
 
-\maketitle
+\begin{flushleft}{
+	\Large
+	\textbf\newline{
+		A note on observation processes in epidemic models
+	}
+}
+\newline
+\\
+Sang Woo Park\textsuperscript{1,2,*}
+Benjamin M. Bolker\textsuperscript{2,3,4}
+\\
+
+\bigskip
+\textbf{1} Department of Ecology and Evolutionary Biology, Princeton University, Princeton, New Jersey, USA
+\\
+\textbf{2} Department of Mathematics and Statistics, McMaster University, Hamilton, Ontario, Canada
+\\
+\textbf{3} Department of Biology, McMaster University, Hamilton, Ontario, Canada
+\\
+\textbf{4} Michael G. DeGroote Institute for Infectious Disease Research, McMaster University, Hamilton, Ontario, Canada
+\\
+\bigskip
+
+*swp2@princeton.edu
+\end{flushleft}
 
@@
 are controlled and are effectively no longer infectious).
-\bmb{Maybe mention King et al. -- completely ignored observation processes}
 
@@
 For example, when $\mathcal R_0 = 5$, such delay in the reporting of new cases can reduce the size of the observed epidemic peak by almost 50\%.
+Note that $i_2(t)$ essentially assumes that the amount of time between when an individual is infected and when an individual reports the infection is exponentially distributed; a fixed delay in reporting time will not change the shape of an epidemic curve.
 
@@
 infection time series underestimates the basic reproduction number and 
-gives a slightly low coverage (Figure 2A). Fitting the infection model to recovery
+gives a slightly low coverage (Figure 2A). 
+Fitting the infection model to recovery
 time series slightly overestimates the basic reproduction number but gives 
-good coverage. Fitting the correct models to their corresponding time series
-gives unbiased estimates and good coverage.
+good coverage. 
+We expect fitting incorrect models to give more biased estimates when $\mathcal R_0$ is higher because the differences in the size of observed epidemic peaks become greater.
+Fitting the correct model gives unbiased estimates and good coverage.
 
@@
 When we assume that the true value of the mean infectious period is known
-and try to estimate the remaining 4 parameters of the SIR model (Figure 2B), fitting the
-incorrect model results in a clearer bias (in opposite directions)
+and try to estimate the remaining 4 parameters of the SIR model (Figure 2B), fitting
+incorrect models results in a clearer bias (in opposite directions)
 and a much lower coverage (cf. \cite{elderd2006uncertainty}).
@@
 
-\bmb{Not including process error. Discuss?}
-\bmb{Constant reporting delays don't matter, if system is autonomous}
-
-Mathematical modeling of infectious disease outbreaks helps us understand 
-how disease spreads in a population; however, epidemic models often make 
-simple assumptions about how cases are reported. 
-We used a deterministic SIR model to show that
-assumptions about observation processes can
-affect the estimates of the basic reproduction number and their associated confidence intervals.
+Mathematical modeling of infectious disease outbreaks helps us understand how disease spreads in a population; 
+however, epidemic models often make simple assumptions about how cases are reported. 
+We used a deterministic SIR model to show that delays in case reports affect the observed shape of an epidemic curve and the estimates of the basic reproduction number.
+Even when the basic reproduction number is small (e.g., $\mathcal R_0 = 2$), fitting incorrect models can introduce small bias and give narrow confidence intervals.
 
@@
 give better coverage probabilities even when wrong observation models are used.
-Nonetheless, misspecifying the observation model can still affect conclusions from 
+Nonetheless, misspecifying the observation model may still affect conclusions from 
 stochastic models and introduce bias.

### 2019-11-27 figure/example.R

@@ -19,7 +19,7 @@ for (i in 1:length(R0vec)) {
     R0=R0vec[i],
     gamma=1,
     N=1e5,
-    rho=0.5
+    rho=1
   )
   
   rr <- run_sir(param=newpar, tmax=200, tlength=0.1)

### 2019-11-27 figure/example.R
tweak

@@ -85,4 +85,4 @@ g3 <- ggplot(fixed_inf_data) +
 gfinal <- arrangeGrob(g1, g2, g3, layout_matrix = matrix(c(1, 1, 2, 3), 2, 2),
                       widths = c(2, 1))
 
-ggsave("example.pdf", gfinal, width=8, height=6)
+ggsave("example.pdf", gfinal, width=9, height=6)

### 2019-11-27 ms/ms.tex
more stuff
.

@@
 
-Note that incidence (the number of \emph{newly} infected individuals) is different from 
-prevalence (the number of \emph{currently} infected individuals) \citep{bjornstad2018epidemics}:
-the observed dynamics of incidence depend on the reporting time step (because the sum of 
-true incidence is equal to the final size of an epidemic), whereas those of
-prevalence do not.
-While they are relatively
-uncommon, some models do not make a clear distinction between prevalence and 
-incidence 
-\citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.
+We emphasize that incidence (i.e., the number of \emph{newly} infected individuals) is different from prevalence (i.e., the number of \emph{currently} infected individuals) \citep{bjornstad2018epidemics}.
+The dynamics of incidence depend on the reporting time step (because the sum of true incidence is equal to the final size of an epidemic), whereas those of prevalence do not.
+We expect the dynamics of incidence and prevalence to be similar only when the reporting time step is similar to the disease generation time \citep{fine1982measles}.
+While they are uncommon, some models do not make a clear distinction between prevalence and 
+incidence \citep{capistran2009parameter, hooker2010parameterizing, yang2013stability, gonzalez2014fractional}.
 
@@
 
-We thank David Earn for providing useful comments.
+We thank David Earn for providing helpful comments on the manuscript. BMB is supported by an NSERC Discovery grant.
 

### 2019-11-27 ms/ms.tex
minor tweaks

@@
 and $\Delta t = 0.1$ time units.
-(B-C) Effects of $\mathcal R_0$ on the relative difference in the size of the epidemic peak ($1 - [\max i_2(t)]/[\max i_1(t)]$) and the difference in the peak timing ($\hat{t}_2 - \hat{t}_1$ where $i_k(\hat{t}_k) = \max i_k(t)$ for $k = 1, 2$) due to lags in reporting time.
-Black circles: simulation results of the deterministic SIR model.
-Black lines: locally estimated scatterplot smoothing (LOESS) curves fitted to simulation results.
+(B-C) Effects of $\mathcal R_0$ on the relative difference in the size of the epidemic peak ($1 - [\max i_2(t)]/[\max i_1(t)]$) and the difference in the peak timing ($\hat{t}_2 - \hat{t}_1$ where $i_k(\hat{t}_k) = \max i_k(t)$ for $k = 1, 2$) due to delays in reporting time.
+Open circles: simulation results of the deterministic SIR model.
+Solid lines: locally estimated scatterplot smoothing (LOESS) curves fitted to simulation results.
 Remaining parameters are held constant ($1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$, and $\Delta t = 0.1$ time units) throughout simulations.
@@
 In order to understand how assumptions about the timing of case reporting affect 
-parameter estimates of the SIR model, we simulate observed epidemic time series
+parameter estimates of the SIR model, we simulate epidemic time series
 (infection time series and recovery time series) 

### 2019-11-27 figure/compare_deterministic.R

@@ -70,10 +70,11 @@ coverdata <- summdata %>%
 g2 <- ggplot(filter(coverdata, inf=="estimated")) +
 	geom_rect(xmin=-Inf, xmax=Inf, ymin=binom.test(95, 100)[[4]][1], ymax=binom.test(95, 100)[[4]][2], alpha=0.1) +
 	geom_hline(yintercept=0.95, lty=2) +
-	geom_point(aes(x=data, y=coverage, col=fit, group=fit), size=5, position = position_dodge(width = 0.5)) +
+	geom_point(aes(x=data, y=coverage, col=fit, group=fit, shape=fit), size=5, position = position_dodge(width = 0.5)) +
 	scale_x_discrete("Simulated timing") +
 	scale_y_continuous("Coverage probability", limits=c(0, 1)) +
 	scale_colour_discrete("Fitted timing") +
+  scale_shape_discrete("Fitted timing") +
 	theme(
 		legend.position=c(0.76, 0.2)
 	)

### 2020-02-13 ms/response.tex
close

\documentclass[12pt]{article}
\usepackage[utf8]{inputenc}

\usepackage{color}

\usepackage{lmodern}
\usepackage{amssymb,amsmath}

\newcommand{\rR}{\mbox{$r$--$\cal R$}}
\newcommand{\RR}{\ensuremath{{\cal R}}}
\newcommand{\RRhat}{\ensuremath{{\hat \cal R}}}
\newcommand{\Rx}[1]{\ensuremath{{\cal R}_{#1}}} 
\newcommand{\Ro}{\Rx{0}}
\newcommand{\Reff}{\Rx{\mathit{eff}}}
\newcommand{\Tc}{\ensuremath{C}}

\newcommand{\rev}{\subsection*}
\newcommand{\revtext}{\textsf}
\setlength{\parskip}{\baselineskip}
\setlength{\parindent}{0em}

\newcommand{\comment}[3]{\textcolor{#1}{\textbf{[#2: }\textsl{#3}\textbf{]}}}
\newcommand{\jd}[1]{\comment{cyan}{JD}{#1}}
\newcommand{\swp}[1]{\comment{magenta}{SWP}{#1}}
\newcommand{\dc}[1]{\comment{blue}{DC}{#1}}
\newcommand{\jsw}[1]{\comment{green}{JSW}{#1}}
\newcommand{\hotcomment}[1]{\comment{red}{HOT}{#1}}

\begin{document}

\noindent Dear Editor:

Thank you for the chance to revise and resubmit our manuscript. 
Below please find our responses to reviewers.

\rev{Reviewer 1}

\revtext{Authors aim to evaluate how using the wrong model in terms of the distributions of the infectious and latent periods affect estimates of $\mathcal R_0$.}

\revtext{Authors did some work in this direction, but I believe more work is needed to address their question. In particular, authors will need to use a model that incorporates realistic distributions of latent and infectious periods. They can accomplish this using the linear chain tricks or alternative integro-differential equation models. Then they can generate simulated data from that "true/realistic" model and go ahead and estimate $\mathcal R_0$ using simple models as they do here.}

\revtext{Having said this, I would like to point out that I am sure there are a number of papers out there that have look at similar questions in the context of different infectious diseases, etc. It'd be good to make sure the relevant literature is considered and appropriately cited.}

We agree that this is an important point. Although our study considers the effect of making incorrect assumptions about the reporting time distributions, these assumptions implicitly depend on latent and infectious period distributions. We have added the following paragraph in the Discussion to address this point:

``Delay distributions in reporting time implicitly depend on latent and infectious 
period distributions. For example, when cases are assumed to be reported
at the end of infectious periods,
the delay distribution is equivalent to the convolution of the latent and
infectious period distributions. These distributions play important roles in 
shaping the epidemiological dynamics, including the probability and size of an outbreak, 
the stability and persistence of a system, and the 
prediction of dynamical transitions. Assumptions about
latent and infectious period distributions can also have large effects on the
estimates of the basic reproductive number.''

\rev{Reviewer 2}

\revtext{This manuscript demonstrates that the timing of when disease cases are reported can influence the performance of statistical models fit to time series of incidence data.  Essentially, the results show that generating data with one reporting scenario and fitting a model that assumes a different reporting scenario leads to sub-optimal inference on the basic reproductive number, $\mathcal R_0$.  In the case when the infectious period is estimated, the degraded inference is not too bad (bias on the order of 5 - 10\%, actual coverage rates of nominally 95\% intervals is off by a few percent).  Coverage rates plummet when the infectious period is assumed fixed, but the bias is still not awful, so it is not clear just how much of a worry the loss of coverage will be for real data, given all the other strong assumptions of the simulation study.}

\revtext{Overall, it seems to me that this is a publishable piece of work.  It may make epidemic modelers think twice about the consequences of their particular assumptions about the process generating incidence data, and such an outcome would propel the field forward.  Technically, the study is soundly executed.  Given the modest bias of the estimates from the mismatched models, I think it's possible that readers may take the reverse message from this study than the one the authors intend.  Namely, readers might interpret these results to indicate that faulty assumptions about the process generating the incidence data are tolerable, especially in light of all the other complications that attend real data, and that have been assumed away here.  But, such a demonstration would be a worthy contribution to the literature as well.}

\revtext{I have one comment that the authors may wish to consider, although this doesn't rise to the level of a critical flaw that must be corrected.  It seemed to me that the Discussion lost sight of the fact that statistical models often have to make simplifying assumptions.  Despite these simulation results, it seems to me that making a simplifying assumption about the process generating the incidence data may not be a terrible thing to do with real data.  When it comes to fitting statistical models to noisy data, models that are "more realistic", in the sense of incorporating more of the underlying mechanism, do not necessarily lead to improved inference.  One has to choose their battles with respect to what one tries to capture in a model, and what one simplifies away.  With real data, it is entirely unclear whether model fit would be improved by fitting a model with a more complicated description of the process generating the incidence data, and it may very well be that such a
question can only be answered on a case-by-case basis.  In the absence of such understanding, it seems to me that the Discussion could strike a somewhat more measured tone regarding the merits of simpler vs. more complicated models.}

We agree. We did not intend to say that modelers should always thrive to build the more complex models. We have revised the final paragraph of the Discussion to better reflect our intention as well the reviewer's comment:

``All statistical models must rely on simplifying assumptions.
Given a range of choices of assumptions one can make, modelers have to
decide which aspect of a system they wish to capture 
at the cost of simplifying other aspects.
However, some assumptions are often more carefully tested than others.
In the case of epidemic models, assumptions about the process of transmission tend to be
the most important ones; processes through which cases are reported are often 
overlooked and simplified.
Our study shows that a seemingly negligible change in the assumptions of an epidemic
model can affect the inference of 
infectious disease transmission.
Although the amount of bias introduced from the misspecification of the observation
model can be small (e.g., less than 5\% in our examples), modelers
trying to make serious predictions may want to control for potential biases.
We caution disease modelers to be more careful about modeling decisions as they may have unexpected consequences on the inference.''

\revtext{As trivia, perhaps I missed something, but the distinction between the boxplots within each pair in the left panels of Fig 2 was not immediately clear to me.  I guessed at the distinction by comparison with the parallel structure of the panels on the right of the figure.}

\revtext{I congratulate the authors on a tidy piece of work, and a well written manuscript.}

\end{document}

### 2020-02-13 ms/ms.tex
close

@@
 \begin{figure}[!t]
-\includegraphics[width=\textwidth]{../figure/example.pdf}
+\includegraphics[width=\textwidth]{example.pdf}
 \caption{
@@
 \begin{figure}
-\includegraphics[width=\textwidth]{../figure/compare_deterministic.pdf}
+\includegraphics[width=\textwidth]{compare_deterministic.pdf}
 \caption{
@@
 
+Delay distributions in reporting time implicitly depend on latent and infectious 
+period distributions. For example, when cases are assumed to be reported
+at the end of infectious periods \citep{breto2009time, he2009plug, lin2016seasonality},
+the delay distribution is equivalent to the convolution of the latent and
+infectious period distributions. These distributions play important roles in 
+shaping the epidemiological dynamics, including the probability and size of an outbreak
+\citep{anderson1980spread}, the stability and persistence of a system
+\citep{keeling1998effect, lloyd2001destabilization, lloyd2001realistic}, and the 
+prediction of dynamical transitions \citep{krylova2013effects}. Assumptions about
+latent and infectious period distributions can also have large effects on the
+estimates of the basic reproductive number \citep{wearing2005appropriate}.
+
 Here, we assumed that the underlying transmission process is deterministic; this 
@@
 
+All statistical models must rely on simplifying assumptions.
+Given a range of choices of assumptions one can make, modelers have to
+decide which aspect of a system they wish to capture 
+at the cost of simplifying other aspects.
+However, some assumptions are often more carefully tested than others.
+In the case of epidemic models, assumptions about the process of transmission tend to be
+the most important ones; processes through which cases are reported are often 
+overlooked and simplified.
 Our study shows that a seemingly negligible change in the assumptions of an epidemic
-model can affect the inference of infectious disease transmission.
-While the amount of bias introduced from the misspecification of the observation
-model can be fairly small (e.g., less than 5\% in our examples), any modelers
-trying to make serious forecasts should be aware of potential biases.
-We caution disease modelers to carefully consider the implications of their model assumptions when developing epidemic models.
+model can affect the inference of 
+infectious disease transmission.
+Although the amount of bias introduced from the misspecification of the observation
+model can be small (e.g., less than 5\% in our examples), modelers
+trying to make serious predictions may want to control for potential biases.
+We caution disease modelers to be more careful about modeling decisions as they may have unexpected consequences on the inference.
 

### 2020-02-13 figure/compare_stochastic.R

@@ -1,74 +1,0 @@
-library(dplyr)
-library(ggplot2); theme_set(theme_bw())
-library(gridExtra)
-
-scale_colour_discrete <- function(...,palette="Dark2") scale_colour_brewer(...,palette=palette)
-scale_fill_discrete <- function(...,palette="Dark2") scale_fill_brewer(...,palette=palette)
-
-load("../data/stochastic.rda")
-
-allpred <- allres <- allsum <- allmle <- list()
-
-load("../analysis/fakedata_incidence.rda")
-
-allpred$incidence <- predres
-allres$incidence <- reslist
-allsum$incidence <- sumlist
-allmle$incidence <- mle
-
-load("../analysis/fakedata_mortality.rda")
-
-allpred$mortality <- predres
-allres$mortality <- reslist
-allsum$mortality <- sumlist
-allmle$mortality <- mle
-
-summdata <- allsum %>%
-  lapply(bind_rows) %>%
-  bind_rows(.id="type") %>%
-  mutate(
-    type=factor(type, levels=c("incidence", "mortality"), 
-                labels=c("measured upon infection", "measured upon recovery"))
-  ) %>%
-  filter(
-    R0 < 2.5
-  )
-
-horizontal <- data.frame(
-  hh=c(max(predict(loess(logLik~R0, data=filter(summdata, type=="measured upon infection")))), 
-       max(predict(loess(logLik~R0, data=filter(summdata, type=="measured upon recovery")))))-qchisq(0.95, 1)/2,
-  type=c("measured upon infection", "measured upon recovery")
-)
-
-mledata <- allmle %>% 
-  lapply(function(x){
-    data.frame(
-      R0=coef(x)[1]
-    )
-  }) %>%
-  bind_rows(.id="type") %>%
-  mutate(
-    type=factor(type, levels=c("incidence", "mortality"), 
-                labels=c("measured upon infection", "measured upon recovery"))
-  )
-
-g1 <- ggplot(summdata) +
-  geom_smooth(aes(R0, logLik), lty=1, se=FALSE, method="loess", span=0.5, col="black", lwd=0.5) +
-  geom_point(aes(R0, logLik), shape=1, size=2) +
-  geom_hline(data=horizontal, aes(yintercept=hh), lty=2) +
-  geom_vline(data=mledata, aes(xintercept=R0)) +
-  geom_rect(data=data.frame(type="measured upon infection"),
[…]

### 2020-02-13 figure/compare_deterministic.R
stuff

@@ -76,17 +76,26 @@ g2 <- ggplot(filter(coverdata, inf=="estimated")) +
 	scale_colour_discrete("Fitted timing") +
   scale_shape_discrete("Fitted timing") +
 	theme(
-		legend.position=c(0.76, 0.2)
+		legend.position="top"
 	)
+
+g_legend<-function(a.gplot){
+  tmp <- ggplot_gtable(ggplot_build(a.gplot))
+  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
+  legend <- tmp$grobs[[leg]]
+  return(legend)
+}
+
+mylegend <- g_legend(g2)
 
 g3 <- g1 %+% filter(summdata, inf=="fixed")
 	
 g4 <- g2 %+% filter(coverdata, inf=="fixed") +
 	theme(legend.position="none")
 
-g5 <- grid.arrange(g1, g2, nrow=1, top=textGrob("A) Estimated infectious period", hjust=1.7))
+g5 <- grid.arrange(g1, g2 + theme(legend.position = "none"), nrow=1, top=textGrob("A) Estimated infectious period", hjust=1.7))
 g6 <- grid.arrange(g3, g4, nrow=1, top=textGrob("B) Fixed infectious period", hjust=2))
 
-gtot <- arrangeGrob(g5, g6, nrow=2)
+gtot <- arrangeGrob(mylegend, arrangeGrob(g5, g6, nrow=2), nrow = 2, heights=c(1, 10))
 
 ggsave("compare_deterministic.pdf", gtot, width=8, height=6)

### 2020-02-13 ms/response.tex
stuff
ok
i0, small wording

@@
 
-We agree that this is an important point.
-Although our study considers the effect of making incorrect assumptions about the reporting time distributions, these assumptions implicitly depend on latent and infectious period distributions.
-We have added the following paragraph in the Discussion to address this point:
+The reviewer may have misconstrued our goals in this paper.
+We were \emph{not} exploring the effects of the process delay distributions (i.e.
+latent and infectious periods), and their mis-specification, on epidemiological dynamics and parameter estimation.
+Rather, we were exploring the effects of reporting delays (and their mis-specification) on parameter estimation.
+
+However, reporting delays do depend implicitly on the process delays. We have added the following paragraph in the Discussion to address this point:
 
@@
 estimates of the basic reproductive number.''
+
+Although we could extend the paper to consider dynamics (and reporting time delays) based on more realistic distributions, our goal here was to evaluate the effects of these assumptions qualitatively, in the simplest possible epidemiological model.
 
@@
 
-We agree.
-We did not intend to say that modelers should always thrive to build the more complex models.
-We have revised the final paragraph of the Discussion to better reflect our intention as well the reviewer's comment:
+We agree. We did not intend to say that modelers should always strive to build the more complex models. We have revised the final paragraph of the Discussion to better reflect our intentions:
 
@@
 
+We have modified the figure to make it clearer.
+
 \revtext{I congratulate the authors on a tidy piece of work, and a well written manuscript.}
 
+Thank you very much!
+
 \end{document}

### 2020-02-13 ms/ms.tex
stuff
i0, small wording

--- 2019-11-27 ms/ms.tex
@@
 A deterministic simulation of the SIR model using the following parameters: 
-$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
+$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i_0 = 1 \times 10^{-4}$,
 and $\Delta t = 0.1$ time units.
@@
 Solid lines: locally estimated scatterplot smoothing (LOESS) curves fitted to simulation results.
-Remaining parameters are held constant ($1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$, and $\Delta t = 0.1$ time units) throughout simulations.
+Remaining parameters are held constant ($1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i_0 = 1 \times 10^{-4}$, and $\Delta t = 0.1$ time units) throughout simulations.
 }
@@
 following parameters:  
-$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i0 = 1 \times 10^{-4}$,
+$\mathcal R_0 = 2$, $1/\gamma = 1$ time units, $N = 1 \times 10^5$, $i_0 = 1 \times 10^{-4}$,
 $\rho = 0.5$, $\theta = 10$, and $\Delta t = 0.1$ time units.
@@
 
+Distributions of reporting time delays implicitly depend on the latent and infectious 
+period distributions. For example, when cases are reported
+at the end of infectious periods \citep{breto2009time, he2009plug, lin2016seasonality},
+the delay distribution is equivalent to the convolution of the latent and
+infectious period distributions. These distributions also play important roles in 
+shaping the epidemiological dynamics,
+including the probability and size of an outbreak
+\citep{anderson1980spread}, the stability and persistence of a system
+\citep{keeling1998effect, lloyd2001destabilization, lloyd2001realistic},
+and the  occurrence of dynamical transitions \citep{krylova2013effects}.
+Assumptions about latent and infectious period distributions can also have large effects on the estimates of the basic reproductive number \citep{wearing2005appropriate}.
+
 Here, we assumed that the underlying transmission process is deterministic; this 
@@
 
+All statistical models rely on simplifying assumptions.
+Given a range of possible assumptions, modelers have to
+decide which aspects of a system to capture 
+at the cost of simplifying other aspects.
+However, some assumptions are often taken for granted.
+Epidemic modelers tend to focus on assumptions about the
+transmission process; processes through which cases are reported are often 
+overlooked and simplified.
 Our study shows that a seemingly negligible change in the assumptions of an epidemic
-model can affect the inference of infectious disease transmission.
-While the amount of bias introduced from the misspecification of the observation
-model can be fairly small (e.g., less than 5\% in our examples), any modelers
-trying to make serious forecasts should be aware of potential biases.
-We caution disease modelers to carefully consider the implications of their model assumptions when developing epidemic models.
+model can affect the inference of 
+infectious disease transmission.
+Although the amount of bias introduced from the misspecification of the observation
+model can be small (e.g., less than 5\% in our examples), modelers
+trying to make serious predictions may want to control for potential biases.
+We caution disease modelers to be more careful about their modeling decisions as they may have unexpected consequences on the inference.
 

### 2020-02-13 ms/response.tex
edits

@@
 
-However, reporting delays do depend implicitly on the process delays. We have added the following paragraph in the Discussion to address this point:
+However, reporting delays do depend implicitly on the process delays. We have added the following paragraph in the Discussion with appropriate citations to address this point:
 
-``Delay distributions in reporting time implicitly depend on latent and infectious 
-period distributions. For example, when cases are assumed to be reported
+``Distributions of reporting time delays implicitly depend on the latent and infectious 
+period distributions. For example, when cases are reported
 at the end of infectious periods,
 the delay distribution is equivalent to the convolution of the latent and
-infectious period distributions. These distributions play important roles in 
-shaping the epidemiological dynamics, including the probability and size of an outbreak, 
-the stability and persistence of a system, and the 
-prediction of dynamical transitions. Assumptions about
-latent and infectious period distributions can also have large effects on the
-estimates of the basic reproductive number.''
+infectious period distributions. These distributions also play important roles in 
+shaping the epidemiological dynamics,
+including the probability and size of an outbreak, the stability and persistence of a system,
+and the  occurrence of dynamical transitions.
+Assumptions about latent and infectious period distributions can also have large effects on the estimates of the basic reproductive number.''
 
@@
 
-``All statistical models must rely on simplifying assumptions.
-Given a range of choices of assumptions one can make, modelers have to
-decide which aspect of a system they wish to capture 
+``All statistical models rely on simplifying assumptions.
+Given a range of possible assumptions, modelers have to
+decide which aspects of a system to capture 
 at the cost of simplifying other aspects.
-However, some assumptions are often more carefully tested than others.
-In the case of epidemic models, assumptions about the process of transmission tend to be
-the most important ones; processes through which cases are reported are often 
+However, some assumptions are often taken for granted.
+Epidemic modelers tend to focus on assumptions about the
+transmission process; processes through which cases are reported are often 
 overlooked and simplified.
@@
 trying to make serious predictions may want to control for potential biases.
-We caution disease modelers to be more careful about modeling decisions as they may have unexpected consequences on the inference.''
+We caution disease modelers to be more careful about their modeling decisions as they may have unexpected consequences on the inference.''
 
